---
title: "Hypothesis Tests Involving Means"
tutorial:
  id: "10-Hypotheses-Means"
output: 
  qpaTutorials::qpa_tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to apply hypothesis tests involving means
---


## Learning Objectives

In this tutorial, you will learn how to conduct hypothesis tests about means. Specifically, we will cover:

* How to use the `t.test()` function to conduct (1) one-sample means tests, (2) difference of means tests involving two independent samples and one variable, and (3) paired t-tests involving a single sample and two variables. 
* Practice walking through the steps for hypothesis testing.

This tutorial assumes that you already are familiar with the logic of hypothesis testing and the specific tests covered here. See the tutorial "Bivariate Description: One Categorical Variable and One Interval Variable" to review how to generate tables of means of an interval variable by categories of a categorical variable and how to compare the distribution of the interval variable based on the values of the categorical variable.


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
library(poliscidata)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
#tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
data("counties", package = "qpaTutorials")
counties$dem2p_percent <- counties$dem2p_vote_share*100
data("df", package = "qpaTutorials")
df$region <- factor(df$region, labels =  c("South", "Northeast", "Midwest", "West"))
library(poliscidata)
```



## Hypothesis Tests Involving Means

There are a variety of hypothesis tests related to the means of interval level variables. All tests involving means can be conducted with the `t.test()` function.

The `t.test()` function performs one and two sample t-tests on a vector of data (a variable). The `t.test()` function takes a number of arguments depending on the type of hypothesis test about means you require. (See ?t.test for the different options.) 

We will illustrate how to use the `t.test()` function to conduct (1) [one-sample means tests]{.important-text}, (2) [difference of means tests]{.important-text} involving two independent samples and one variable, and (3) [paired t-tests]{.important-text} involving a single sample and two variables. In each case, we will present examples that pose a research question and walk through the five steps of hypothesis testing to draw a conclusion about our question.


## One-sample means tests 

A [one-sample means test]{.important-text} is used to test whether the mean of a single variable $Y$ in a single sample is equal to a hypothesized value. We use a one-sample means test to answer questions like "Is the mean value of presidential approval in our sample equal to a particular value, say 42%?"

### Example 1: Vote for Hillary Clinton in US Counties
 
To illustrate this test, consider the variable **dem2p_percent**, the percent of the vote for Hillary Clinton in US counties. The variable is in the data frame **counties**.  Hillary Clinton won 48.2\% of the popular vote overall.  We will investigate whether the average percent of the vote she won in US counties is equal to this percentage or not equal to it.

**Step 1. Formulate a null hypothesis**.


```{r quiz-null-hypothesis-clintonvote, echo=FALSE}
question("Which of the following is the appropriate null hypothesis?",
  answer("$H_0: \\bar{Y}=48.2$", message="The null hypothesis refers to the population parameter, not the sample statistic"),
  answer("$H_0: \\mu_{Y}=48.2$", correct=TRUE, message="The null hypothesis refers to the population statistic and is set equal to the hypothesized value. "),
  allow_retry = TRUE
  )
```

**Step 2. Formulate an alternative hypothesis**.

```{r quiz-alternative-hypothesis-clintonvote, echo=FALSE}
question("Which of the following statements about the alternative hypothesis are true?",
  answer("$H_A: \\mu_Y=48.2$", message="$H_A: \\mu_Y=48.2$ is the null hypothesis"),
  answer("The alternative hypothesis is two-tailed", correct=TRUE, message="A two-tailed alternative hypothesis specifies that the true population value is $\\ne$ to the hypothesized value given by the null hypothesis."),
  answer("$H_A: \\mu_Y\\ne48.2$", correct=TRUE, message="Because we want to test whether the mean percent of the vote is different from that in the population, the alternative hypothesis is two-tailed and thus $\\ne$ to 48.2."),
  answer("The alternative hypothesis is one-tailed", message="A one-tailed alternative hypothesis would specify that the true population mean is either > or < 48.2, rather $\\ne$ as we specified above"),
  allow_retry = TRUE,
  try_again = "Hint: There are two correct statements."
  )
```

**Step 3. Compute the appropriate test statistic.** 

The test statistic we use to test the significance of a sample mean, denoted $z$ is given by:

$$z=\frac{\bar{Y} - \mu_0}{\sigma/\sqrt{n}}$$
The z-score tells us how many standard deviations away from the mean an observation is. The further away our sample z-score is from the null hypothesis value, the further out in the tails of the distribution the statistic is and the more confident we can be that the null hypothesis is wrong. However, we don't typically know the population standard deviation ($\sigma$). This means we have to switch things up a bit. We replace $\sigma$ with $s$ (the estimated standard deviation) and calculate a $t$-score:

$$t=\frac{\bar{Y} - \mu_0}{s/\sqrt{n}}.$$


The t-statistic will be further from zero the more $\bar{Y}$ differs from $\mu_0$ and the smaller the standard error (Recall $s.e.=s/\sqrt{n}$).  Note that you do not need to remember the formula for the test statistics. You should understand the logic of each test statistic, but R will compute them for you.

To compute a one-sample $t$-test in R, we use the `t.test()` function. The function requires we pass it the name of the variable for which to calculate the sample mean and the argument `mu`, which gives the value under the null hypothesis. Running the following code will produce the value of the t-test, the corresponding p-value, a confidence interval for the mean, and the sample mean.

By default, R computes the test statistic assuming the alternative hypothesis is two-tailed (the null is $\ne$ to the hypothesized value). This means you do not have to specify the test is two-tailed. If we wish to test the null hypothesis against a one-tailed alternative -- that the mean is higher or lower than the hypothesized value under the null -- we add the `alternative` argument, setting it to `"greater"` if the mean is hypothesized to be greater than under the null and `"less"` if the mean in is hypothesized to be smaller than the null.

The code below conducts the two-tailed one-sample means test to test our hypothesis.


**Try it: Run one-sample t-test on Clinton vote share**

```{r exercise-ttest-clintonvote, exercise = TRUE, exercise.cap="Interactive exercise: Run the code to conduct a one-sample t-test comparing the mean dem2p_percent to a hypothesized value. Use the Run Code button to execute and interpret the test output. No hints are available."}
t.test(counties$dem2p_percent, mu = 48.2)
```
A lot of information is included in the output for the test. The output begins by telling us that the test conducted is a one-sample t-test and that the data used for the test is **counties$dem2p_percent**. This confirms we computed the test we desired. The t-statistic is reported to be -51.294. This is the information of central interest to us in step 3. The output also reports the degrees of freedom for the test, or "df." This tells us how many observations were used to calculate the test minus the number of parameters estimated by the test (one, the mean of **dem2p_percent**). The next piece of information is the p-value. We'll skip that for now.  We can see that the alternative hypothesis is two-sided because the output reports the mean is "not equal" to 48.2. The output reports the 95% confidence interval. Recall that the confidence interval gives the lower and upper limit on the range of values around our sample statistic that mark the interval that is likely to contain the unknown population parameter with 95% confidence. Finally, we see the sample "mean of x," the sample statistic, or mean value of **dem2p_percent**, in the sample: about 33.34%. 

**Step 4. Translate the test statistic into a p-value.**

R reports the two-tailed p-value associated with the t-statistic in scientific notation as "p-value < 2.23-16."  This is 2.23 to the negative 16th or 0.000000000000000223. When the p-value is this small, we report p<.001. **Never report more than three decimals in conjunction with a p-value.**

**Step 5. Draw a conclusion.**

```{r quiz-ttest-interpretation, echo=FALSE}
question("Which of the following statements are true?",
  answer("Because the p-value is less than 0.05, we can reject the null hypothesis with 95% confidence and therefore conclude the mean Democratic share of the vote in counties is different from 48.2", correct=TRUE),
  answer("Because the p-value is less than 0.05, we cannot reject the null hypothesis with 95% confidence and conclude the mean Democratic share of the vote in counties is not different from 48.2",  message="Remember if p is low the null must go!"),
  answer("Because the absolute value of the t-statistic is greater than 1.96, we can reject the null hypothesis with 95% confidence and conclude the mean Democratic share of the vote in counties is different from 48.2", correct=TRUE),
  answer("Because the confidence interval does not include the null value, we can reject the null hypothesis with 95% confidence and conclude the mean Democratic share of the vote in counties is different from 48.2", correct=TRUE),
  answer("There is less than a one percent probability the null hypothesis is true so we conclude that the mean Democratic share of the vote in counties is different from 48.2", correct=TRUE, message="If the p-value is less than 0.05, the t-statistic will be greater than 1.96 (for a two-tailed test), and the confidence interval will not include the value under the null (here 48.2). Each of these facts means we can reject the null hypothesis with 95% confidence. The p-value gives the probability we observe the mean this far from the hypothesized value if in fact the null is true. Since the p-value is less than 0.01, there is less than a one percent probability the null is true."),
  allow_retry = TRUE,
  try_again = "Hint: Make sure to select all true statements."
  )
```

### Example 2: Proportion of white residents in US counties


We might also be interested to know whether, in the average county in the US in 2016, the proportion of whites was the same as the country as a whole, 0.763. The variable **prop_white** in the data frame **counties** contains the proportion of Blacks in a given county. Note that the variable is measured as a proportion, it is not a categorical variable.


**Step 1. Formulate a null hypothesis** and **Step 2. Formulate the alternative hypothesis**


```{r quiz-ttest-hypothesis-proportion-white, echo=FALSE}
question("Which of the following is the appropriate null and alternative hypothesis?",
  answer("$H_0: \\bar{Y}=0.763, H_A: \\bar{Y}\\ne0.763$", message="The null and alternative hypothesis refer to the population parameter, not the sample statistic"),
  answer("$H_0: \\mu_{Y}\\ne0.763, H_A: \\mu_{Y}=0.763$", message="The null hypothesis is set equal to the value under the hypothesis to be tested and the alternative is not equal"),
  answer("$H_0: \\mu_{Y}=0.763, H_A: \\mu_{Y}\\ne0.763$", correct=TRUE),
  allow_retry = TRUE,
  try_again = "Try again."
  )
```

**Step 3. Compute the appropriate test statistic.** 

Write the code to test this hypothesis.

**Practice: Test the mean of proportion of White population against national proportion**
```{r exercise-ttest-hypothesis-proportion-white, exercise = TRUE, exercise.cap="Interactive exercise: Write code to test whether the mean prop_white in counties differs from the national proportion of 0.763 using a hypothesis test. Use the Run Code button to execute and click Submit Answer to check correctness. 2 hints are available."}

```

```{r exercise-ttest-hypothesis-proportion-white-hint-1}
# Hint 1 of 2
You are comparing a sample mean to a specific value.

Think about which function in R is used for a one-sample test of a mean.
```

```{r exercise-ttest-hypothesis-proportion-white-hint-2}
# Hint 2 of 2
Use the variable prop_white from the counties data.

Inside the function, specify the null value of 0.763 
using the argument for the hypothesized mean.
```
     
```{r exercise-ttest-hypothesis-proportion-white-solution}
t.test(counties$prop_white, mu = 0.763)
```


```{r exercise-ttest-hypothesis-proportion-white-check}
grade_code("You got it!")
```

**Step 4. Translate the test statistic into a p-value.** 


```{r quiz-pvalue-proportion-white, echo = FALSE}
question_text(
  "Enter the p-value you would report from the test output. Begin your answer with p (not p-value) and include no spaces in your answer:",
  answer("p<2.2e-16", message = "Do not report scientific notation or more than three decimals."),
  answer("p<.001", correct = TRUE, message="Never report more than three decimal points. A zero before the decimal point is not necessary."),
  answer("p<0.001", correct = TRUE, message="Never report more than three decimal points. A zero before the decimal point is not necessary."),
   allow_retry = TRUE,
  try_again = "Try again."
)
```



**Step 5. Draw a conclusion.**



```{r quiz-pvalue-interpretation, echo=FALSE}
question("Which of the following statements are true?",
  answer("Because the p-value is less than 0.05, we can reject the null hypothesis with 95% confidence and conclude that the mean proportion of whites in counties is different from 0.763", correct=TRUE),
  answer("Because the absolute value of the t-statistic is greater than 1.96, we can reject the null hypothesis with 95% confidence and conclude that the mean proportion of whites in counties is different from 0.763", correct=TRUE),
  answer("Because the confidence interval does not include the null value, we can reject the null hypothesis with 95% confidence and conclude that the mean proportion of whites in counties is different from 0.763", correct=TRUE),
  answer("There is less than a one percent probability the null hypothesis is true so we conclude that the mean proportion of whites in counties is different from 0.763", correct=TRUE, message="If the p-value is less than 0.05, the t-statistic will be greater than 1.96 (for a two-tailed test), and the confidence interval will not include the value under the null (here 0.763). Each of these facts means we can reject the null hypothesis with 95% confidence. The p-value gives the probability we observe the mean this far from the hypothesized value if in fact the null is true. Since the p-value is less than 0.01, there is less than a one percent probability the null is true. Thus, we conclude that the mean proportion of whites in counties is different from 0.763"),
  answer("Because the p-value is less than 0.05, we cannot reject the null hypothesis",  message="Remember if p is low the null must go!"),
  allow_retry = TRUE,
  try_again = "Hint: Make sure to select all true statements."
  )
```



## Difference of means test: Two independent samples, one variable

A [difference of means test]{.important-text} involves a comparison of the means in two groups. We focus here on the difference of means in two independent samples. This means that there is no overlap in the membership of cases in each sample. We use the difference of means test to test whether the means in two groups are the same in the population based on the sample data.

### Example 1: Proportion of Black residents in US counties won by Hillary Clinton and Donald Trump


We might wish to test whether the proportion of Blacks in a county was **higher** in counties that Hillary Clinton won than in those counties Donald Trump won.


**Step 1. Formulate a null hypothesis** and **Step 2. Formulate the alternative hypothesis**

```{r quiz-hypothesis-proportion-black-clintonvote, echo=FALSE}
question("Which of the following correctly states the null and alternative hypothesis?",
  answer("$H_0: \\mu_C=\\mu_T$ and $H_A: \\mu_C\\ne\\mu_T$", message="The alternative hypothesis specified above is a one-tailed test"),
  answer("$H_0: \\bar{Y}_C=\\bar{Y}_T$ and $H_A: \\bar{Y}_C>\\bar{Y}_T$", message="The null and alternative hypothesis do not apply to the sample statistic but to the population parameter."),
  answer("$H_0: \\mu_C=\\mu_T$ and $H_A: \\mu_C>\\mu_T$", correct=TRUE, message="Our alternative hypothesis is that counties Clinton won were **more** likely to have a higher proportion of Blacks than those Trump won so the alternative hypothesis is one-tailed with  $\\mu_C>\\mu_T$."),
  answer("$H_0: \\mu_C=\\mu_T$ and $H_A: \\mu_C<\\mu_T$",  message="The hypothesis stated above is one-tailed, but we expect the proportion of Blacks to be larger in counties Hillary Clinton won"),
  allow_retry = TRUE
)
```



**Step 3. Compute the appropriate test statistic.** 


The test statistic we use to test the difference of means is the Welch 2-sample t-test:

$$ t=\frac{\bar{Y}_1 - \bar{Y}_2-0}{\sqrt{\frac{s^2_1}{n_1}+\frac{s^2_2}{n_2}}}$$

where 1 and 2 denote the two groups, here Trump majority of Clinton majority. We assume the two groups are independent, i,e., that no one is in both groups.


There are two ways to test whether the mean value of a single variable, $Y$, in R is the same in two different populations (in two groups) against the alternative that it is different.  Which we use depends on how many categories are counted by the categorical variable. If, like here, there are exactly two values the categorical variable can take, we can use what R calls a formula. We list the name of the (interval level) variable for which we want to calculate the mean, then we type a tilde (~) and then the name of the variable that defines the two groups: `t.test(df$IntervalVariable ~ df$CategoricalVariable)`. If the categorical variable contains three or more categories, then you *cannot* use the formula notation. Instead, you must use indexing. (You can also use indexing if the categorical variable contains just two categories.) To test this hypothesis using indexing you specify `t.test(df$IntervalVariable[df$CategoricalVariable == category1],df$IntervalVariable[df$CategoricalVariable == category2]`.

If we wish to test the null hypothesis against a one-tailed alternative -- that the mean is higher or lower in one of the two groups -- we add the `alternative` argument, setting it to `"greater"` if the mean in group zero is hypothesized to be greater  than in group one and `"less"` if the mean in group zero is hypothesized to be smaller than in group one The default is a two-tailed alternative hypothesis, so you don't need to specify this argument if the alternative hypothesis is two-tailed.


The variable **TrumpMajority** (coded 0 if Hillary Clinton won a majority of the county's vote and 1 if Donald Trump won a majority) and **prop_black** (the proportion of the county residents who are Black) in the **counties** data frame contain the information needed to test the hypothesis that the mean proportion of Black's in counties won by Hillary Clinton is higher than that in counties won by Donald Trump.

Note that we specify `alternative = "greater"` to test this hypothesis because we expect the mean in group zero to be higher than the mean in group one. Anytime you use a one-tailed test, take care to ensure you choose the appropriate alternative given the coding of the variables! I've computed this test using both formulations of the test.

**Try it: Compare counties' mean proportion of Black population by TrumpMajority**

```{r exercise-ttest-proportion-black-trumpmajority, exercise = TRUE, exercise.cap="Interactive exercise: Run the code to conduct a one-tailed t-test comparing mean prop_black between counties by TrumpMajority. Use the Run Code button to execute and interpret the results. No hints are available."}
t.test(counties$prop_black ~ counties$TrumpMajority, alternative = "greater")
```

**Try it: Compare counties' mean proportion of Black population by TrumpMajority (two-sample)**
```{r exercise-ttest-twosample-trumpmajority, exercise = TRUE, exercise.cap="Interactive exercise: Run the code to conduct a one-tailed two-sample t-test comparing prop_black between counties where Clinton and Trump won using indexing. Use the Run Code button to execute and interpret the results. No hints are available."}
t.test(counties$prop_black[counties$TrumpMajority == 0], counties$prop_black[counties$TrumpMajority == 1], alternative = "greater")
```


```{r quiz-proportion-black-trumpmajority, echo=FALSE}
question("Which of the following statements are true?",
  answer("The mean proportion of Blacks in counties Clinton won was 0.23", correct = TRUE, message="Group 0 refers to the first response option for TrumpMajority, which corresponds to a Clinton majority."),
  answer("The mean proportion of Blacks in counties Trump won was 0.23", correct = FALSE, message="Group 1 refers to the second response option for TrumpMajority, which corresponds to a Trump majority. The average proportion of Blacks in counties Trump won is about 0.075."),
  answer("The counties that Clinton won had, on average, about  0.23 - 0.07 = 0.16 greater proportion that were Black than did counties Trump won.", correct = TRUE,  message="0.23 - 0.07 = 0.16 is approximately the difference in the mean proportion black in counties Clinton won, group 0, and counties Trump won, group 1."),
  allow_retry = TRUE,
  try_again = "Hint: Make sure to select all true statements."
  )
```

**Step 4. Translate the test statistic into a p-value.**



```{r quiz-pvalue-black-trumpmajority, echo = FALSE}
question_text(
  "Enter the p-value you would report from the test output. Begin your answer with p (not p-value) and include no spaces in your answer:",
  answer("p<2.2e-16", message = "Do not report scientific notation or more than three decimals."),
  answer("p<.001", correct = TRUE, message="Never report more than three decimal points. A zero before the decimal point is not necessary."),
  answer("p<0.001", correct = TRUE, message="Never report more than three decimal points. A zero before the decimal point is not necessary."),
   allow_retry = TRUE,
  try_again = "Try again."
)
```

**Step 5. Draw a conclusion.**



```{r quiz-interpret-pvalue-black-trumpmajority, echo=FALSE}
question("Which of the following statements are true?",
  answer("Because the p-value is less than 0.05, we cannot reject the null hypothesis with 95% confidence", message="Remember p is low the null must go"),
  answer("Because the confidence interval does not include the null value, we cannot reject the null hypothesis", message="If a confidence interval does not include the value under the null hypothesis, we can reject the null hypothesis. "),
answer("There is less than a 1% chance that we would observe a difference in the group means this large (and positive) if, in fact, the  means were equal or the proportion Black was higher in counties Trump won. ", correct=TRUE, message="If the p-value is less than 0.05, the 95% confidence interval will not include the value under the null. Each of these facts means we can reject the null hypothesis with 95% confidence. The p-value gives the probability we observe the mean this far from the hypothesized value if, in fact, the null is true. Since the p-value is less than 0.01, there is less than a one percent probability the null is true. "),
  allow_retry = TRUE,
  try_again = " "
  )
```

```{r quiz-conclusion-black-trumpmajority, echo=FALSE}
question("What substantive conclusion can we draw from this hypothesis test?",
  answer("The proportion of Blacks in counties won by Hillary Clinton is significantly lower than that in counties Donald Trump won", message="Group 0 is comprised of counties Clinton won, therefore the proportion is significantly higher in counties won by Hillary Clinton than in counties won by Donald Trump."),
    answer("The proportion of Blacks in counties won by Hillary Clinton is significantly higher than that in counties Donald Trump won.", correct=TRUE, message=""),
    answer("The mean proportion Black in counties Clinton won is not greater than the proportion Black in counties won by Donald Trump for an alpha level of .05.", message = "This would be our conclusion if we failed to reject the null hypothesis."),
  allow_retry = TRUE,
  try_again = " "
  )
```

### Example 2: GDP and democracy around the world 

In the data frame **world**, the variable **democ_regime** takes the value "No" if a country is not a democracy and "Yes" if it is. How do we test the null hypothesis that the GDP (**gdp08**) is different among countries with democratic and non-democratic regimes?

**Step 1. Formulate a null hypothesis** and **Step 2. Formulate the alternative hypothesis**


```{r quiz-hypothesis-democracy-gdp, echo=FALSE}
question("Which of the following correctly states the null and alternative hypothesis?",
  answer("$H_0: \\mu_D=\\mu_{ND}$ and $H_A: \\mu_{D}>\\mu_{ND}$", message = "This is a one-tailed alternative hypothesis. The alternative hypothesis stated above is two-tailed, "),
  answer("$H_0: \\mu_D=\\mu_{ND}$ and $H_A: \\mu_{D}\\ne\\mu_{ND}$", correct=TRUE, message="The alternative hypothesis is that the means are different and thus is two-tailed such that we specify $H_A: \\mu_{D}\\ne\\mu_{ND}$"),
  allow_retry = TRUE
)
```

**Step 3. Compute the appropriate test statistic.** 

Try to produce the t-test without looking at the hint. You do not need to use indexing in this example.

**Practice: Compare mean GDP by democratic regime status**

```{r exercise-ttest-democracy-gdp, exercise = TRUE, exercise.cap="Interactive exercise: Write code to conduct a two-sample t-test comparing mean gdp08 between democratic and non-democratic countries using formula notation. Use the Run Code button to execute and click Submit Answer to check correctness. 2 hints are available."}

```

```{r exercise-ttest-democracy-gdp-hint-1}
# Hint 1 of 2
You are testing whether the mean of one variable differs across two groups.

Think about which variable is the outcome and which variable defines the groups.
```

```{r exercise-ttest-democracy-gdp-hint-2}
# Hint 2 of 2
Use gdp08 as the outcome variable and democ_regime as the grouping variable.

Write them inside t.test() using formula notation with a tilde (~).
```


```{r exercise-ttest-democracy-gdp-solution}
t.test(world$gdp08 ~ world$democ_regime)
```


```{r exercise-ttest-democracy-gdp-check}
grade_code("You got this!")
```


```{r quiz-ttest-democracy-gdp, echo=FALSE}
question("Which of the following statements are true?",
  answer("The mean value of GDP in 2008 among democracies is 585.5484", correct = TRUE, message="Group Yes refers to democracies, group No to nondemocracies. 585.5484 is the mean for group Yes."),
  answer("Democracies had a higher mean GDP in 2008 than did non democracies", correct = TRUE, message="The difference in the mean of GDP in 2008 among democracies (group Yes) is greater than the mean among nondemocracies (group No)."),
  answer("The mean value of GDP in 2008 among democracies is 226.0468",  message="Group Yes refers to democracies, group no to nondemocracies. 226.0468 is the mean for group no."),
  allow_retry = TRUE,
  try_again = " "
  )
```



## Paired t-test: one-sample means test, two variables 
 
A [paired t-test]{.important-text} matches two variables in one sample and tests whether the average difference between values on these two variables is zero in the population against the alternative that it is not zero. 

To illustrate this test, we will consider whether the Gini coefficient of countries around the world changed from 2008 to 2010. The Gini coefficient is a measure of income equality. It ranges from 0, indicating perfect equality (where everyone receives an equal share of income), to 1, perfect inequality (where only one recipient or group of recipients receives all the income).  Thus this test allows us to determine whether, on average, income inequality around the world increased, decreased, or stayed the same. The data frame **world** contains the variables **gini08** and **gini10**. 

Let's test whether the GINI coefficients increased, on average, from 2008 to 2010.


**Step 1. Formulate a null hypothesis** and **Step 2. Formulate the alternative hypothesis**


```{r quiz-hypothesis-gini, echo=FALSE}
question("Which of the following correctly states the null and alternative hypothesis?",
  answer("$H_0: \\mu_{gini10}=\\mu_{gini08}$ and $H_A: \\mu_{gini10}>\\mu_{gini08}$", message = "This null and alternative hypothesis specifies a difference of means hypothesis not a mean difference hypothesis but we want to test whether, on average, the mean difference, the mean of (**gini10**-**gini08**), is zero or greater than zero."),
  answer("$H_0: \\mu_{gini10-gini08}=0$ and $H_A: \\mu_{gini10-gini08}>0$", correct=TRUE, message="We want to test whether, on average, **gini10**-**gini08** is equal to zero or greater than zero."),
  allow_retry = TRUE
)
```

**Step 3. Compute the appropriate test statistic.** 

The test statistic for the paired samples $t$-test, denoted $t$, follows the same formula as the one sample $t$-test.

$$t=\frac{\bar{Y}_{diff}-0}{se_{\bar{Y}}}$$

where

$$se_{\bar{Y}} = \frac{s_{diff}}{\sqrt{n}}$$

and

+ $\bar{Y}_{diff}$ = Sample mean of the differences;
+ $n$ = Sample size (i.e., number of observations);
+ $s_{diff}$ = Sample standard deviation of the differences;
+ $se_{\bar{Y}}$ = Estimated standard error of the mean.


To conduct a paired t-test in R, use `t.test(X1, X2, paired=TRUE)`. If we wish to test the null hypothesis against a one-tailed alternative -- that the mean difference is higher or lower than the value under the null hypothesis (0), we add the `alternative` argument, setting it to `"greater"` if the mean difference is hypothesized to be greater  than the hypothesized value and `"less"` if the mean difference is hypothesized to be smaller than the hypothesized value.

Write the code to compute the test statistic. Be sure to list **gini10** first if you specify `"greater"` for the `alternative` argument, as the first value (**gini10**) is then greater than the second (**gini08**) under the alternative hypothesis. If you list **gini08** first, then you need to specify `"less"` for the `alternative` argument. Both specifications will produce the same result. For the exercise below, list **gini10** first.


**Practice: Run paired t-test on Gini coefficients**
```{r exercise-ttest-gini, exercise = TRUE, exercise.cap="Interactive exercise: Write code to conduct a paired t-test comparing gini10 and gini08, specifying the correct alternative hypothesis. Use the Run Code button to execute and click Submit Answer to check correctness. 2 hints are available."}

```

```{r exercise-ttest-gini-hint-1}
# Hint 1 of 2
You are comparing the same countries at two time points,
so think about which type of t-test 
accounts for paired observations.

Also consider what it means for 
one measure to be greater than the other.
```

```{r exercise-ttest-gini-hint-2}
# Hint 2 of 2
Use the two Gini variables from the world data inside t.test(),
set the argument that indicates pairing,

and specify a one-sided alternative where 
the first variable is expected to be larger.
```

```{r exercise-ttest-gini-solution}
t.test(world$gini10, world$gini08, paired = TRUE, alternative = "greater")
```

```{r exercise-ttest-gini-check}
grade_code("You are getting the hang of this")
```

```{r quiz-differences-gini, echo=FALSE}
question("Which of the following statements are true?",
  answer("The average difference in countries' Gini coefficients in 2010 and 2008 of about -0.54 indicates that on average inequality grew from 2008 to 2010 by 0.54.", message="The mean difference is calculated as the mean of gini10 minus gini08, this implies that the 2008 value is higher (there is more inequality in 2008) than the 2010 value."),
  answer("The average difference in countries' Gini coefficients in 2010 and 2008 of about -0.54 indicates that on average inequality fell from 2008 to 2010 by 0.54.", correct=TRUE, message="The mean difference is calculated as the mean of gini10 minus gini08, this implies that the 2008 value is higher (there is more inequality in 2008) than the 2010 value."),
  allow_retry = TRUE,
  try_again = " "
  )
```



**Step 4. Translate the test statistic into a p-value** and **Step 5. Draw a conclusion.**

```{r quiz-pvalue-gini, echo=FALSE}
question("Which of the following statements are true?",
  answer("Because the p-value is greater than 0.05, we can reject the null hypothesis with 95% confidence", message="The p-value is not greater than 0.05"),
  answer("Because the confidence interval includes zero, we cannot reject the null hypothesis with 95% confidence", correct=TRUE, message="When the 95% confidence interval includes the value under the null (here 0), we cannot reject the null hypothesis at the 0.05 level."),
  answer("There is a 93.12 percent chance the mean of the difference between the two Gini coefficients is 0.5396694", message="It is not true that the p-value tells us the probability that the estimated mean difference is true. Instead, it tells us that there is a 93.12 percent chance we'd observe the mean of the difference between the two Gini coefficients as great as we observe if the null hypothesis is true."),
  allow_retry = TRUE,
  try_again = " "
  )
```

```{r quiz-interpret-pvalue-gini, echo=FALSE}
question("What substantive conclusion can we draw from this hypothesis test?",
  answer("There is no statistical evidence that the levels of inequality changed significantly, on average, from 2008 to 2010",  message="We did not test this hypothesis. Instead, we tested whether the average difference in GINI coefficients (gini10 - gini08) was greater than zero"),
  answer("There is statistical evidence that the levels of inequality were significantly higher, on average, in 2010 than 2008",  message="Because we fail to reject the null hypothesis that the mean difference in the GINI coefficient (gini10 - gini08) is greater than zero, this statement is false"),
    answer("There is no statistical evidence that the levels of inequality were significantly higher, on average, in 2010 than 2008",  correct = TRUE, message="Because we fail to reject the null hypothesis that the mean difference in the GINI coefficient (gini10 - gini08) is greater than zero, this statement is true"),
      answer("We can be 95% confident that the levels of inequality were significantly higher, on average, in 2008 than 2010",   message="Even though levels of inequality were, on average, higher in 2008, because we fail to reject the null hypothesis that the mean difference in the GINI coefficient (gini10 - gini08) is greater than zero, this statement is false"),
  allow_retry = TRUE,
  try_again = " "
  )
```


## The Takeaways

Often we will be interest in comparing means to a given population value, or comparing means in two groups of cases, or comparing means of a single population on some measure at multiple time points. We do so with means hypothesis tests. We use the R function `t.test()` to conduct hypothesis tests regarding means. 

+ A [one sample means test]{.important-text} is used to test whether the mean of a single variable $Y$ in a single sample is equal to a hypothesized value. If we are comparing a sample mean to a specific population value we specify the name of the variable and the argument `mu`, where `mu` is set equal to the population value: `t.test(counties$dem2p_percent, mu=48.2)`.
+ If we want to test whether the mean in two groups is the same we conduct a [difference of means test]{.important-text}:
    + If the groups are identified by another variable, the first argument is the variable for which we wish to calculate the means and we specify the tilde followed by the name of the variable that denotes the categories: `t.test(df$IntervalVariable ~ df$CategoricalVariable)`, where **df** is the name of your dataframe.  
    + If the categorical variable contains more than two values use: `t.test(df$IntervalVariable[df$CategoricalVariable == category1],df$IntervalVariable[df$CategoricalVariable == category2]`. category1 and category2 should be the values of the categories you wish to compare. 
    + We can only use the difference of means test to compare means in two groups. If the nominal variable has more than two categories, you may need to create a new, two-caetgory variable to use in the test.
+ If we want to compare the mean difference between two variables in a single sample, we conduct a [paired t-test]{.important-text}, which matches two variables in one sample and tests whether the average difference between values on these two variables is zero in the population against the alternative that it is not zero: `t.test(world$gini10, world$gini08)`. But note that we cannot compare the mean difference of just any two variables, the comparison must be logical. We cannot, for example, compare the mean of the difference between the age of a survey respondent and their level of education. That makes no sense! Typically, a paired t-test will involve the comparison of two variables that measure the same concept, measured identically, at two points in time.

By default, all tests are two-tailed, but we can specify `alternative = "greater"` or `alternative = "less"` to identify the form of the alternative hypothesis.

The steps involved in hypothesis testing are always the same:

1. Formulate a null hypothesis. 
2. Formulate an alternative hypothesis.  
3. Compute the appropriate test statistic.
4. Translate the test statistic into a p-value.
5. Draw a conclusion.

Remember that the null and alternative hypotheses are about the population parameter(s). In the case of means tests, the appropriate test statistic is the t-test, but the formula used to calculate the test statistic varies depending on which of the types of means test we use.

```{r tutorial-success-banner-10, results="asis", echo=FALSE}
cat('
<p role="status"
   aria-label="Excellent! You have completed this tutorial and can now conduct hypothesis tests involving means."
   style="font-size: 28px; font-weight: 600; color: #2e7d32; text-align: center; margin-top: 50px;">
  <span aria-hidden="true">🏆 Excellent work! 🏆</span>
</p>
')
```


