---
title: "Bivariate Description: One Categorical (Nominal or Ordinal) Variable and One Interval Variable"
tutorial:
  id: "08-Bivariate-Description-Categorical-Interval-Data"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to describe the relationship between a categorical and an interval level variable.
---

## Learning Objectives

In this tutorial, you will continue learning how to describe the relationship between two variables. This time we will focus on how to do so when one of the variables is categorical (either nominal or ordinal) and the other variable is interval level. Specifically, we will cover:

* How to conduct a descriptive analysis of the relationship between nominal or ordinal variables and an interval variable by comparing means across categories.
* How to use the functions in the R package <span style="color:DarkGreen">dplyr</span> to produce means by categories of a nominal/ordinal variable.
* How to plot means using the `ggplot()` function in the R package <span style="color:DarkGreen">ggplot2</span>.


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
library(poliscidata)
library(descr)
library(ggplot2)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
counties <- qpaTutorials::counties
counties$mortality_risk_25_45_change <- counties$mortality_risk_25_45_change*100
counties$employ_pop_ratio_25_64 <- counties$employ_pop_ratio_25_64*100
counties$wage_growth <- counties$wage_growth*100
counties$dem2p_percent <- counties$dem2p_vote_share*100
counties$rural <-1
counties$rural[counties$rural_urban<=3] <- 0
counties$rural[counties$rural_urban>=8] <-2

fHouse <- qpaTutorials::fHouse
sum(is.na(fHouse$FHStatus))
#fHouse <-filter(fHouse, FHStatus==)
fHouse$FHStatus2[fHouse$FHStatus=="NF"] <- 0
fHouse$FHStatus2[fHouse$FHStatus=="PF"] <- 1
fHouse$FHStatus2[fHouse$FHStatus=="F"] <- 2
fHouse$FHStatus2[fHouse$FHStatus=="NA"] <- NA

fHouse$FHStatus<-fHouse$FHStatus2
#fHouse$FHStatus <- factor(fHouse$FHStatus, ordered=TRUE, levels = c("NF", "PF", "F"))
world <- world
```


## Bivariate Description

As we discussed in the previous lesson, bivariate description gives us the first look at  the relationship between two variables. We introduced cross-tabs and grouped bar plots to describe the relationship between variables measured at the nominal and/or ordinal level. In this tutorial, we look at the tools available for describing the relationship between one nominal or ordinal variable and an interval variable.

If our hypothesis involves the relationship between an interval variable and a nominal or ordinal variable, we compare the relationship between them by comparing the mean of interval variable for each value of the nominal/categorical (or ordinal) variable.

## Comparing Means

To motivate our analysis, consider the following hypothesis: <span style="color:chocolate"> In a comparison of [counties], those having [lower wage growth] will be more likely to  [give a majority of their support to Trump] than will those having [lower wage growth].</span> The causal explanation is that lower wage growth leads to dissatisfaction with the incumbent party, here the Democrat, and thus that counties with lower wage growth will vote against the Democrat, Hillary Clinton.

The data frame **counties** contains the interval level variable **wage_growth**, which gives the rate at which wages in a county have grown over the last four years (as a percent) and the nominal variable **TrumpMajority**, which takes a value of 0 if Hillary Clinton won a majority of the county's votes and 1 if Donald Trump won a majority. Our hypothesis leads us to expect that the mean rate of wage growth should be lower in counties Trump won.


We return to the functions in the  <span style="color:DarkGreen">dplyr package</span> for this analysis. Recall:

1. The `group_by()` function breaks the data down based on the values of a variable. 
2. The `summarise()` function calculates a summary statistic for the data passed to it.


### Mean Wage Growth by Victor

To calculate the mean of **wage_growth** for counties won by Clinton and Trump, we take the data frame **counties** and pipe it to the `group_by()` function where the grouping variable is **TrumpMajority** and then pipe the result to  the `summarize()` function to calculate the means of **wage_growth** for counties casting a majority of votes for Clinton and for Trump. We will call our new variable **mean** and set it equal to `mean(wage_growth, na.rm=TRUE)`, removing any missing values by setting the `na.rm` argument to `TRUE` to avoid getting NA for the means if any counties have missing values. I've also summarized the frequencies in each category using the `n()` function inside `summarise()`. A warning message will be produced but can be ignored.

```{r meanwage, exercise=TRUE}
library(dplyr)
counties %>%
  group_by(TrumpMajority)%>%
  summarise(freq = n(), mean = mean(wage_growth, na.rm = TRUE))
```

Here we see that the mean rate of wage for counties won by Clinton is 4.86 while that for counties won by Trump is 4.23, consistent with our hypothesis.


### Mean Mortality Risk by Victor

We might also believe that counties with worse health outcomes punish the incumbent party. The data frame **counties** also contains the variable **mortality_risk_25_45_change**, which gives the change in the probability of dying in this particular age range (in percent), given that one is in that age range and has made it thus far, compared to the same probability four years ago. We thus might hypothesize: <span style="color:chocolate"> In a comparison of [counties], those having [higher mortality risk among adults 45-45 than 4 years ago] will be more likely to  [give a majority of their support to Trump] than will those having [lower mortality risk among adults 45-45].</span>

Does the evidence support this hypothesis? If counties with increasing mortality rates are more likely to cast a majority of votes for Donald Trump, what should we see?

We would expect the mean change in mortality risk (**mortality_risk_25_45_change**) for counties that cast a majority of votes for Donald Trump would be greater than in those counties that cast a majority of votes for Hillary Clinton.  Edit the XXXX in the code below to produce the mean of the change in mortality risk for counties won by Clinton and Trump.

```{r meanrisk, exercise=TRUE}
counties %>%
  XXXX(XXXX)%>%
  XXXX(freq = XXXX,mean = mean(XXXX))
```

```{r meanrisk-hint-1}
Remember to group_by() TrumpMajority and use summarise()
to calculate frequencies with the n() function and means.
Don't forget to include the na.rm argument and set it to
TRUE.
Remember to separate arguments with a comma.
```

```{r meanrisk-solution}
counties %>%
  group_by(TrumpMajority)%>%
  summarise(freq = n(), mean = mean(mortality_risk_25_45_change, na.rm = TRUE))
```
```{r meanrisk-check}
grade_code()
```


What have we learned? Counties that cast a majority of votes for Donald Trump on average experienced an *increase* of 1.87 in their mortality risk (their risk of dying between the ages of 25 and 45, provided they were in that age group 4 years prior) while counties that cast a majority of their votes for Hillary Clinton, on average, saw a decrease in their mortality risk of 2.05. This suggests counties that were worse off in terms of their health outcomes were more likely to cast a majority of votes for Donald Trump, providing support for the hypothesis.

### Mean GPD by Regime Type

As a final example, consider the hypothesis: <span style="color:chocolate"> In a comparison of [countries], those that [are more democratic] will be more likely to  [have better economic outcomes] than will those that are [less democratic].</span> 

Calculate the means of the variable **gdp08**,
GDP in billions in 2008 (World Bank) by **dem_level4**,
Regime type (Economist 2014), in the data frame **world**. Try not to look at the hints! Name the counts `freq` and the means `mean`.


```{r meangdp, exercise=TRUE}

```

```{r meangdp-hint-1}
Pipe the world data frame to group_by() dem_level4 and
use summarise() to calculate frequencies and the means
of gdp08.
```

```{r meangdp-hint-2}
Make sure you separate arguments with a comma and
use the na.rm argument (set to TRUE)
```

```{r meangdp-solution}
world %>%
  group_by(dem_level4)%>%
  summarise(freq = n(), mean = mean(gdp08, na.rm = TRUE))
```

```{r meangdp-check}
grade_code()
```

The table of means tell us that the mean is highest, 1394.33, among full democracies ("Full Democ"), followed by 354.79 among partial democracies ("Part Democ"), as we would expect, but authoritarian governments have the next highest means at 320.43, and finally, hybrid democracies have the lowest GDP.

## Plotting Means by Group 

We can plot the means as well. To do so, we will pipe the results from our tables above to `ggplot()` so that it becomes the data for the plot. We can then use the variables in our table in our plot. The basic steps of plotting remain the same:

1.  Tell `ggplot()` the data that contains the variable(s) to plot. (Here simply include `ggplot()` in the sequence of pipes after generating the means.)
2.  Tell `ggplot()` which variable(s) to plot and on which axis. We do this by mapping the variables to the appropriate axis using the `mapping` argument and the `aes()` function.
3.  Tell `ggplot()` how to plot the data by adding a **geom** layer. To plot the means we need a new geom layer. We want a simple point for each mean. Thus we use `geom_point()`.
4.  Add layers that make the plot more informative by adding labels and formatting scales.


Let's start with a basic plot of the means of **wage_growth** by **TrumpMajority** in the data frame **counties** without using informative labels or other visual attributes that make it easier to interpret. The first lines of code are identical to those used above to produce the table of means. (Name the counts `freq` and the means `mean`.) See if you can edit them without looking above or at the solution.

Notice that I've loaded the <span style="color:DarkGreen">ggplot2</span> package with the `library()` function.


```{r meanwageplot, exercise=TRUE}
library(ggplot2)
counties %>%
  group_by(XXXX)%>%
  summarise(XXXX) %>%
  ggplot(mapping = aes(x = factor(TrumpMajority), y = mean) ) +
    geom_point() 
```


```{r meanwageplot-hint}
We want to group by TrumpMajority and calculate
the mean. We need to call the mean *mean* to use it
in the ggplot function.
```

```{r meanwageplot-solution}
library(ggplot2)
counties %>%
  group_by(TrumpMajority)%>%
  summarise(freq = n(), mean = mean(wage_growth, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = factor(TrumpMajority), y = mean) ) +
    geom_point() 
```

```{r meanwageplot-check}
grade_code()
```


Now let's improve on the plot by adding a layer of labels using the `labs()` function and replacing "0" and "1" on the x-axis with "Clinton" and "Trump" using the `scale_x_discrete()` function. 

```{r meanwageplot2, exercise=TRUE}
counties %>%
  group_by(TrumpMajority)%>%
  summarise(freq = n(), mean = mean(wage_growth, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = factor(TrumpMajority), y = mean)) +
    geom_point() +
  labs(title = "Mean Change in Wage Growth Among Counties \nCasting a Majority of Votes for Clinton and Trump",
       y = "Wage Growth", 
       x = NULL, 
       caption = "Source: Linn, Nagler, and Zilinsky") +
  scale_x_discrete(labels = c("Clinton", "Trump"))
```

Finally, let's change the `size` and `shape` of the points inside `geom_point()` and change the limits of the y axis by adding another layer to the plot with `scale_y_continuous()`. This function tells `ggplot()` to rescale a continuous y axis. Inside we can use the `limits` argument to set the lower and upper limits. We list the lower limit followed by a comma and the upper limit inside the `c()` function: 0 to 10 seems a good choice.


Look carefully at the code to make sure you understand what each argument is doing!

```{r meanwageplot3, exercise=TRUE}
counties %>%
  group_by(TrumpMajority)%>%
  summarise(freq = n(), mean = mean(wage_growth, na.rm = TRUE)) %>%
ggplot(mapping = aes(x = factor(TrumpMajority), y = mean)) +
    geom_point(size = 3, shape = 19) +
  labs(title = "Mean Change in Wage Growth Among Counties \nCasting a Majority of Votes for Clinton and Trump",
       y = "Change in Mortality Risk", 
       x = NULL, 
       caption = "Source: Linn, Nagler, and Zilinsky") +
  scale_x_discrete(labels = c("Clinton", "Trump")) +
  scale_y_continuous(limits = c(0, 10)) 
```

Try to complete the code below to plot the mean **mortality_risk_25_45_change** for counties Trump won and Clinton won. When generating means name the counts `freq` and the means `mean`. Use "Mean Change in Mortality Risk Among Counties \\nCasting a Majority of Votes for Clinton and Trump"" for the title of the plot and label the y-axis "Change in Mortality Risk". Specify the caption as "Source: Linn, Nagler, and Zilinsky". Label the x-axis tick marks "Clinton" and "Trump" and omit the x-axis title. Set the y-axis limits to -10 to 10.

```{r meanrisk2, exercise=TRUE}
counties %>%
  group_by()
```

```{r meanrisk2-hint-1}
Remember to group_by() TrumpMajority and use summarise() to calculate
frequencies and means. Don't forget to pipe the results to ggplot().
```

```{r meanrisk2-hint-2}
Remember that if you pipe the data to ggplot you do not
need to specify the data argument.
The first argument is thus the mapping argument, which 
should be set to aes. Inside this function set the x
argument to factor(TrumpMajority) and the y argument 
to mean.

Make sure to balance your parentheses!
End the line with a +
```

```{r meanrisk2-hint-3}
The geom we want is point, so specify geom_point
You can specify a size and shape argument to control the
look of the points.
End with a plus.
```

```{r meanrisk2-hint-4}
Add the labs function and specify the title, y, x, and
caption arguments (separate them with commas).
```

```{r meanrisk2-hint-5}
Fix the scales with scale_x_discrete setting the labels 
argument to c("Clinton", "Trump") and
set the limits argument c(-10, 10) inside the
scale_y_discrete function.

Don't forget the +!
```

```{r meanrisk2-solution}
counties %>%
  group_by(TrumpMajority)%>%
  summarise(freq = n(), mean = mean(mortality_risk_25_45_change, na.rm = TRUE)) %>%
ggplot(mapping = aes(x = factor(TrumpMajority), y = mean)) +
    geom_point(size = 3, shape = 19) +
  labs(title = "Mean Change in Mortality Risk Among Counties \nCasting a Majority of Votes for Clinton and Trump",
       y = "Change in Mortality Risk", 
       x = NULL, 
       caption = "Source: Linn, Nagler, and Zilinsky") +
  scale_x_discrete(labels = c("Clinton", "Trump")) +
  scale_y_continuous(limits = c(-10, 10)) 
```

```{r meanrisk2-check}
grade_code()
```

Finally, try to produce the plot of mean GDP, **gdp08**, by the level of democracy, **dem_level4** from the **world** data frame. Since **dem_level4** is character class, you don't need to wrap it in the `factor()` function. Set the title for the plot to "Mean GDP by Level of Democracy", label the y-axis "Mean GDP", and do not use a label for the x-axis (set it to NULL).  When generating means name the counts `freq` and the means `mean`. You do not need to adjust the y-axis limits. Label the x-axis tick marks (in order) "Full Democracy", "Partial Democracy", "Hybrid", and "Authoritarian". Omit the x-axis title.

```{r meangdp2, exercise=TRUE}

```

```{r meangdp2-hint-1}
Remember to group_by() dem_level4 and use summarise() to calculate frequencies and means.
Don't forget to pipe your results to gpplot().
```

```{r meangdp2-hint-2}
Did you set x=dem_level and y=mean?
Did you add geom_point?
Did you add labs()?
```

```{r meangdp2-solution}
world %>%
  group_by(dem_level4)%>%
  summarise(freq = n(), mean = mean(gdp08, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = dem_level4, y = mean)) +
  geom_point() +
  labs(title = "Mean GDP by Level of Democracy",
       x = NULL,
       y = "Mean GDP") +
  scale_x_discrete(labels = c("Full \nDemocracy", "Partial \nDemocracy", "Hybrid", "Autoritarian"))
```

```{r meangdp2-check}
grade_code()
```


## Visualizing the distributions of an interval level variable based on categories of a nominal or ordinal variable

We might want to compare the distribution of some interval variable categories of our nominal or ordinal variable. This allows us to see not just how the means compare but also to look at the dispersion and shape of the distribution across categories. For example, we might want to examine  **mortality_risk_25_45_change**  by **TrumpMajority** to see how similar the distribution of mortality risk is in counties Clinton won compare to those Trump won. 

### Comparing distributions with `geom_histogram()`

You know how to produce a histogram. We place the interval level variable on the x-axis and use `geom_historgram()` to plot the data. If we want a separate histogram for each value of **TrumpMajority** we assign that variable to `fill`. Below I've added a new layer to the plot using `facet_wrap()`. This function will create separate panels in our plot based on the values of another variable. Here we want separate panels for each value of **TrumpMajority** so we pass this variable to `facet_wrap()`. Note that we need to include a tilde (~) before the variable name inside the function.

```{r hist, exercise=TRUE}
ggplot(data = counties, mapping = aes(x = mortality_risk_25_45_change, fill = factor(TrumpMajority))) +
  geom_histogram()  +
  labs(title = "Distribution Change in Mortality Risk by Candidate \nReceiving a Majority of Votes Cast", 
       x = "Change in Mortality Risk", 
       y = "Frequency",
       caption = "Source: Linn, Nagler, and Zilinksy")  +
    facet_wrap(~factor(TrumpMajority)) +
  scale_fill_discrete(name = "",
                      labels = c("Clinton", "Trump"))
```


### Comparing distributions with `geom_boxplot()`

Often it is useful to present box plots rather than histograms. This allows a direct comparison of the median value of the interval variable across categories of the nominal or ordinal variable, as well as the dispersion in the data. To illustrate we will generate box plots of **lifeex_total**,  life expectancy at birth, for each category of the nominal variable **dem_level4** in the data frame **world**.

Three things are different from the histogram plot above. First, we replace `geoom_histogram()` with `geom_boxplot()`. Next, we don't need to use `facet_wrap()`. Finally, we don't need the legend, so we add `guides(fill=FALSE)` in another layer. This tells `ggplot()` that the legend (or guide) associated with the `fill` aesthetic is not to be included.

```{r box, exercise=TRUE}
ggplot(data = world, mapping = aes(x = dem_level4, y = lifeex_total, fill = dem_level4)) +
  geom_boxplot()  +
  labs(title = "Distribution of Life Expectancy at Birth", 
       x = "Democracy Level", 
       y = "Age")  +
  scale_x_discrete(labels = c("Full Democracy", "Partial Democracy", "Hybrid", "Authoritarian")) +
guides(fill = FALSE)
```

Here we can readily see that median life expectancy is highest among full democracies and that the range of life expectancy in this group of countries is also tightly clustered. In contrast, authoritarian countries have a lower median life expectancy and a wider range of them.  We can also tell that among partial democracies there are several outlier countries with unusually short life expectancies.

## Takeaways

The key things to remember about examining the relationship between a nominal or ordinal variable and an interval level variable are:

1. We can compare means of the interval level variable across categories of a nominal or ordinal variable using `group_by()` and `summarise()` in the R package <span style="color:DarkGreen">dplyr</span>.
2. We can plot means using `ggplot()` in the <span style="color:DarkGreen">ggplot2</span> package by piping the means to `ggplot()` and using `geom_point()`.
3. We can compare the distribution of an interval variable across categories of a nominal or ordinal variable in two ways:
    + Using `geom_histogram()` and specifying the nominal/ordinal variable as the `fill` aesthetic. `facet_wrap()` can be used to place the histograms in separate panels.
    + Using `geom_boxplot()` and specifying the nominal/ordinal variable as the `fill` aesthetic.
