---
title: "R Basics: Part 1"
tutorial:
  id: "01-R-Part1"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn the basics of how R works.
---

<style> /* Only show focus indicators when actually focused */
button:focus-visible, .btn:focus-visible {
  outline: 2px solid #6c757d;
  outline-offset: 1px;
}
/* Remove default focus styles that might conflict but preserve cursor */
button:focus, .btn:focus {
  outline: none;
}
/* Ensure proper cursor states for buttons */
button, .btn {
  cursor: pointer;
}
button:disabled, .btn:disabled {
  cursor: not-allowed;
}
/* Ensure keyboard users can see focus on code editors */
.ace_editor.ace_focus {
  border: 2px solid #6c757d !important;
}
/* Better contrast for exercise areas */
.tutorial-exercise {
  border: 1px solid #f0f0f0;
  border-radius: 4px;
  padding: 15px;
  margin: 20px 0;
}
/* Improved font and readability */
body {
  font-size: 16px;
  line-height: 1.5;
}
/* Accessible replacement for blue text */
.important-note {
  background-color: #e3f2fd;
  border-left: 4px solid #1976d2;
  padding: 12px;
  margin: 15px 0;
  font-weight: bold;
  border-radius: 4px;
}
/* Success/completion messages */
.success-note {
  background-color: #e8f5e8;
  border-left: 4px solid #4caf50;
  padding: 12px;
  margin: 15px 0;
  font-weight: bold;
  border-radius: 4px;
}
/* Package name styling */
.package-name {
  background-color: #e8f5e8;
  color: #2e7d32;
  padding: 2px 4px;
  border-radius: 3px;
  font-weight: bold;
}
/* Fix for Run Code button hover states - more specific targeting */
.tutorial-exercise-actions .btn {
  cursor: pointer !important;
}
.tutorial-exercise-actions .btn:hover {
  cursor: pointer !important;
}
.tutorial-exercise-actions .btn:disabled {
  cursor: not-allowed !important;
}
/* Override any conflicting learnr cursor rules */ [data-label] .btn,
.tutorial-exercise .btn,
button[data-label] {
  cursor: pointer !important;
}
/* Accessible table styling */
.table-container {
  overflow-x: auto;
  margin: 20px 0;
}
table {
  width: 100%;
  border-collapse: collapse;
  border: 1px solid #ddd;
}
th, td {
  border: 1px solid #ddd;
  padding: 12px;
  text-align: left;
}
th {
  background-color: #f5f5f5;
  font-weight: bold;
}
caption {
  padding: 10px;
  text-align: left;
  font-size: 1.1em;
  margin-bottom: 10px;
}
/* Focus styles for table cells */
th:focus, td:focus {
  outline: 2px solid #6c757d;
  outline-offset: 1px;
}
/* Skip navigation for keyboard users */
.skip-nav {
  position: absolute;
  top: -40px;
  left: 6px;
  background: #000;
  color: #fff;
  padding: 8px 16px;
  text-decoration: none;
  z-index: 1000;
  border-radius: 4px;
  font-weight: bold;
}
.skip-nav:focus {
  top: 6px;
  outline: 2px solid #fff;
}
/* Enhanced focus visibility for tutorial navigation */
.tutorial-panel-name:focus,
.tutorial-panel-name:focus-visible {
  outline: 1px solid #6c757d !important;
  outline-offset: 1px !important;
  background-color: rgba(108, 117, 125, 0.1) !important;
}
/* Better focus for exercise containers */
.tutorial-exercise:focus-within {
  outline: 1px solid #6c757d;
  outline-offset: 1px;
  background-color: rgba(108, 117, 125, 0.05);
}
/*New*/ /* Enhanced error message accessibility */
.alert-danger, .error, .tutorial-exercise-error {
  border-left: 4px solid #d32f2f;
  padding: 12px;
  margin: 10px 0;
  background-color: #ffebee;
  border-radius: 4px;
}
.alert-success, .tutorial-exercise-success {
  border-left: 4px solid #4caf50;
  padding: 12px;
  margin: 10px 0;
  background-color: #e8f5e8;
  border-radius: 4px;
}
/* Ensure error messages are visible and well-contrasted */
.tutorial-exercise-error {
  color: #d32f2f;
  font-weight: bold;
}
.tutorial-exercise-success {
  color: #2e7d32;
  font-weight: bold;
}
/* Focus styles for elements with error associations */
.ace_editor[aria-describedby]:focus {
  border-color: #d32f2f !important;
  box-shadow: 0 0 0 2px rgba(211, 47, 47, 0.2) !important;
}
</style>

<script>
document.addEventListener('DOMContentLoaded', function() {
  
  // Simple, reliable keyboard shortcut help
  document.addEventListener('keydown', function(e) {
    if (e.ctrlKey && e.shiftKey && e.key === '?') {
      e.preventDefault();
      alert(`R Tutorial Navigation Help:

KEYBOARD LIMITATIONS:
- Tab only moves between headings (not buttons or code boxes)
- Enter after Tab will jump to that heading
- Interactive exercises require mouse/trackpad clicks
- Code editing requires clicking in code boxes first

WHAT WORKS WITH KEYBOARD:
- Ctrl+F to search within the tutorial
- Tab to navigate between section headings
- Screen readers can access all text content

IMPORTANT: This tutorial platform has limited keyboard accessibility.
Interactive elements like "Run Code" and "Submit Answer" buttons
require mouse/trackpad interaction.`);
    }
  });
  
  // Screen Reader Instructions - Collapsible accessibility guidance
  function addScreenReaderInstructions() {
    const instructionsDiv = document.createElement('div');
    instructionsDiv.setAttribute('role', 'region');
    instructionsDiv.setAttribute('aria-label', 'Accessibility Instructions');
    instructionsDiv.innerHTML = `
      <div style="background: #f8f9fa; border: 1px solid #dee2e6; border-radius: 6px; margin: 15px 0;">
        <button id="accessibility-toggle" style="width: 100%; padding: 12px; background: none; border: none; text-align: left; cursor: pointer; font-weight: bold; color: #495057;" 
                aria-expanded="false" aria-controls="accessibility-content">
          ▶ Accessibility & Navigation Guide (Click to expand)
        </button>
        
        <div id="accessibility-content" style="display: none; padding: 0 15px 15px 15px;">
          <div style="margin-bottom: 15px;">
            <h4 style="color: #6c757d; font-size: 16px; margin-bottom: 8px;">Keyboard Navigation:</h4>
            <ul style="margin: 0; padding-left: 20px; line-height: 1.6; font-size: 14px;">
              <li><strong>Press Ctrl+Shift+?</strong> for detailed keyboard help dialog</li>
              <li>Tab moves between section headings only</li>
              <li>Enter after Tab jumps to that heading</li>
              <li>Ctrl+F to search within tutorial</li>
            </ul>
          </div>
          
          <div style="margin-bottom: 15px;">
            <h4 style="color: #6c757d; font-size: 16px; margin-bottom: 8px;">For Screen Reader Users:</h4>
            <ul style="margin: 0; padding-left: 20px; line-height: 1.6; font-size: 14px;">
              <li>All text content is accessible to screen readers</li>
              <li>Interactive code exercises require mouse interaction</li>
              <li>Success messages will be announced as they appear</li>
            </ul>
          </div>
          
          <div style="margin-bottom: 15px;">
            <h4 style="color: #6c757d; font-size: 16px; margin-bottom: 8px;">Platform Limitations:</h4>
            <ul style="margin: 0; padding-left: 20px; line-height: 1.6; font-size: 14px;">
              <li>Exercise buttons require mouse clicks</li>
              <li>Code editing needs mouse interaction</li>
              <li>Interactive elements cannot be reached via Tab</li>
            </ul>
          </div>
          
          <div style="margin-bottom: 15px;">
            <h4 style="color: #6c757d; font-size: 16px; margin-bottom: 8px;">Recommended Approach:</h4>
            <ul style="margin: 0; padding-left: 20px; line-height: 1.6; font-size: 14px;">
              <li>Read content using screen reader navigation</li>
              <li>Use heading navigation between sections</li>
              <li>Consider working with assistance for interactive exercises</li>
            </ul>
          </div>
          
          <p style="margin: 0; font-size: 13px; color: #6c757d; font-style: italic;">
            This platform has inherent accessibility limitations. We've enhanced what's possible while being transparent about constraints.
          </p>
        </div>
      </div>
    `;
    
    // Add click handler for toggle
    setTimeout(function() {
      const toggleBtn = document.getElementById('accessibility-toggle');
      const content = document.getElementById('accessibility-content');
      
      if (toggleBtn && content) {
        toggleBtn.addEventListener('click', function() {
          const isExpanded = content.style.display !== 'none';
          content.style.display = isExpanded ? 'none' : 'block';
          toggleBtn.setAttribute('aria-expanded', !isExpanded);
          toggleBtn.innerHTML = (isExpanded ? '▶' : '▼') + ' Accessibility & Navigation Guide (Click to ' + (isExpanded ? 'expand' : 'collapse') + ')';
        });
      }
    }, 500);
    
    // Try multiple ways to find the right insertion point
    let insertLocation = null;
    
    // Try 1: Look for the h2 with "Introduction" text 
    const headings = document.querySelectorAll('h2');
    for (let i = 0; i < headings.length; i++) {
      if (headings[i].textContent.includes('Introduction')) {
        insertLocation = headings[i];
        break;
      }
    }
    
    // Try 2: Just use the first h2 if Introduction not found
    if (!insertLocation) {
      insertLocation = document.querySelector('h2');
    }
    
    // Insert the accessibility guide immediately after the heading
    if (insertLocation) {
      insertLocation.parentNode.insertBefore(instructionsDiv, insertLocation.nextSibling);
      console.log('Accessibility guide added successfully');
    } else {
      console.log('Could not find insertion point for accessibility guide');
    }
  }
  
  // Enhance success messages with exercise context
  setInterval(function() {
    const successMessages = document.querySelectorAll('.alert-success:not([data-enhanced])');
    successMessages.forEach(function(msg) {
      msg.setAttribute('data-enhanced', 'true');
      msg.setAttribute('role', 'status');
      msg.setAttribute('aria-live', 'polite');
      
      // Add exercise context to success messages
      const exerciseContainer = msg.closest('.tutorial-exercise');
      if (exerciseContainer) {
        const originalText = msg.textContent;
        if (!originalText.includes('Exercise completed')) {
          msg.textContent = 'Exercise completed successfully: ' + originalText;
        }
      }
    });
  }, 2000);
  
  // Enhance error messages for screen readers
setInterval(function() {
  const errorMessages = document.querySelectorAll('.alert-danger:not([data-enhanced]), .tutorial-exercise-error:not([data-enhanced])');
  errorMessages.forEach(function(msg) {
    msg.setAttribute('data-enhanced', 'true');
    msg.setAttribute('role', 'alert');
    msg.setAttribute('aria-live', 'assertive');
  });
}, 2000);

  // Run the accessibility guide function
  setTimeout(addScreenReaderInstructions, 1000);
  
});
</script>


## Introduction {#main-content}

This is the first in a series of tutorials you will work through over the course
of the semester. The tutorials are designed to teach you the ins and outs of
data analysis in R. This tutorial is designed to familiarize you with some of
the basic principles of using the statistical software package R and provide
background knowledge that will underlie the code -- the instructions the
software uses to extract some information from the data -- we write later in the
semester.

The tutorials will typically begin by introducing code with examples that you
will run (using the "Run Code" button). This will be followed by additional
examples in which you complete or edit code and submit your work to be checked
(using the "Submit Answer" button).  Finally, you will be asked to produce code
from scratch.  A "Hint" or "Solution" button will appear to help you if you are
stuck.  Don't be afraid to try out code before you look at hints.
<div class="important-note"> <strong>Important:</strong> You will make mistakes.
R, like any coding language, requires meticulous attention to things like
matching parentheses, including commas in the right places, and spelling
functions, arguments, and object names correctly. When you encounter errors --
and you will do so often -- always look for these types of mistakes first!
</div>

You should work through the tutorials beginning to end, completing each
exercise. If you run out of time, you can exit and return to the tutorial later.
If you are running the tutorials on a local installation of R Studio, your
progress will be saved when you return.

I encourage you to resist the temptation to view the solution and copy it.
Typing out the code will help you to understand what each exercise is doing and
to recall code in your own work. Take notes as you work through the tutorials. A
good strategy would be to create a list of the functions introduced, note the
name of the tutorial that introduced them, and write a short description. This
will provide a help sheet to reference when you need to write your own code from
scratch in class, for problem sets, or for your final project.

## Learning Objectives

This tutorial is designed to familiarize you with some of the basics of using R.
Specifically, you will learn:

* How to use R as a calculator
* How to assign a value or values to an object with <-
* How to use functions
* How to determine the class of an object with the `class()` function
* How to create and use vectors using the `c()` function
* How to select and replace elements of a vector using indexing
* How to install packages with the `install.packages()` function and access the functions in them with the `library()` function

```{r setup, include=FALSE}
library(learnr)
#library(tidyverse)
library(knitr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
#tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488, 0.537, 0.520, 0.511, 0.513, 0.493)
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry","Obama","Obama", "Clinton", "Biden", "Harris")
dem.candidates <- c(first.6.candidates, last.6.candidates)
num.list<-c(1,2,3,4,5)
```

## R as a calculator {#exercises}

It is easy to use R like a calculator. Here's an example of adding 2 numbers.
You can add/subtract, multiply/divide, or raise a number to a power using `+`,
`-`, `*`, `/`, `^`.

**Explore: Basic arithmetic in R**
Try running the code below, then change the numbers and operators to practice
different calculations. Use the "Run Code" button to execute your code. Note
that R ignores spaces, so 14 + 2000 will produce the same output as 14+2000.

```{r calculator-practice, exercise=TRUE, exercise.cap="Interactive exercise: Practice basic arithmetic operations in R. Use the Run Code button to execute the code and see results."}
14 + 2000
```

The `[1]` you see with the line of output is just R telling us that we have just
one number produced by the code. Later we will see cases where a command we run
produces an output that contains multiple numbers (or words). Each line of
output will begin with some number in square brackets [ ], with the number
denoting the index value of the output that starts that line.

## Objects

**R can store information as an object with a name of your choice. Once you have created an R object, you can refer to it by name.**

Picking a name for your object:

* Choose an informative name
* Object names cannot start with a number
* Object names cannot contain spaces
* Object names cannot include special characters that have specific meanings in R. This includes $ and # and %.

The Global Environment window in R Studio will show a list of all the objects
you create.

**We use the assignment operator `<-` to assign some value to an object.**

**Try it: Create your first object**
Let's assign the sum of 5+3 to an object called **my.first.object**. Use the
"Run Code" button to execute the following code and see how assignment works.

```{r objects-create-first, exercise=TRUE, exercise.cap="Interactive exercise: Create and assign a value to your first R object using the assignment operator. Use the Run Code button to execute."}
my.first.object <- 5 + 3
```

Notice that the result is not displayed in this case. We can see the value of an
object only by asking R to show it to us. Simply enter the object name.

**Try it: Display an object's value**
Use the "Run Code" button to see how to display what's stored in an object:

```{r objects-print-value, exercise=TRUE, exercise.cap="Interactive exercise: Display the value stored in an R object by typing the object name. Use the Run Code button to execute."}
my.first.object <- 5 + 3
my.first.object
```

If we assign a different value to the same object name, the value of the object
will change.

**Try it: Reassign an object**
Use the "Run Code" button to see what happens when you assign a new value to an
existing object:

```{r objects-reassign, exercise=TRUE, exercise.cap="Interactive exercise: Practice reassigning a new value to an existing object and observe how the value changes. Use the Run Code button to execute."}
my.first.object <- 50 + 30
my.first.object
```

Object names are case sensitive: **my.first.object** is different from
**MY.FIRST.OBJECT**. Try asking R to tell you the value of **MY.FIRST.OBJECT**.
What happens?

**Try it: Test case sensitivity**
Use the "Run Code" button to see what error occurs with incorrect
capitalization:

```{r objects-case-error, exercise=TRUE, exercise.cap="Interactive exercise: Observe error message from incorrect object name capitalization. This demonstrates R's case sensitivity. Use the Run Code button to execute and see the error."}
MY.FIRST.OBJECT
```

You will see error messages a lot! It's part of the learning process. The trick
is to learn how to fix them! In this case, the error message is pretty clear.
Looking in the Global Environment in RStudio you would see that
**MY.FIRST.OBJECT** does not exist.

###  Storing strings (characters/words etc.) as objects

To assign a string of characters to an object, we use quotation marks. Spaces
ARE allowed.

**Try it: Create a character object**
See how to store text as an object using quotation marks:

```{r objects-create-string, exercise=TRUE,exercise.cap="Interactive exercise: Create an object containing character text using quotation marks. Use the Run Code button to execute and see the error."}
this.class <- "PLSC 309"
this.class
```

R treats numbers like characters if we tell it to do so by enclosing them in
quotes:

**Try it: Numbers as characters**
Observe what happens when you put numbers in quotation marks. Use the "Run Code"
button to execute the code:

```{r objects-number-as-string, exercise=TRUE, exercise.cap="Interactive exercise: Store numbers as character text using quotation marks. Use the Run Code button to see how numbers become text when quoted."}
another.object <- "100"
another.object
```

If you refer to an object that is a character string in an arithmetic operation,
you will see an error message. See what happens if you try to divide "100" by
10:

**Try it: Arithmetic error with character strings**
Use the "Run Code" button to see the error when trying arithmetic with character
strings:

```{r objects-string-arithmetic-error, exercise=TRUE, exercise.cap="Interactive exercise: Observe error when attempting arithmetic operations on character strings. Use the Run Code button to see the error message that occurs."}
another.object <- "100"
another.object / 10
```

## Functions

Functions are very important in R. They can be used to perform a variety of
tasks. **A function takes one or more inputs and returns some output.**
For example, the `sqrt()` function takes as its input some number or numeric
object for which we want to know the square root. The output is the square root
of that number or object. In R, the syntax is generally `funcname(input)`, where
`funcname` is the function name and `input` is one or more *arguments*. Some
functions take no arguments, some take one argument, and others take several
arguments. When there is more than one argument, they are separated by a comma.
Note that omitting a required comma and having unbalanced parentheses are
frequent sources of error.

The function `sqrt()` takes one argument, named `x`, which is the number or
object for which we want the square root.

**Practice: Calculate a square root**
Try taking the square root of 4. Enter your code in the box below and click "Submit Answer" to check if your code is correct. If you get stuck, use the "Hint" button for guidance. There are 2 progressive hints available, with the final hint showing the complete solution.

```{r functions-square-root, exercise=TRUE, exercise.cap="Interactive exercise: Calculate the square root of 4 using the sqrt function. Enter your code and click Submit Answer to check correctness. Two hints available."}

```

```{r functions-square-root-hint-1}
# Hint 1 of 2: Function Structure
# In R, functions follow the pattern: function_name(input)
# For square root, you need the sqrt() function with a number inside the parentheses.
```

```{r functions-square-root-hint-2}
# Hint 2 of 2: Specific Syntax  
# Type: sqrt(4)
# This calls the sqrt function with 4 as the input argument.
```

```{r functions-square-root-solution}
# Complete Solution
sqrt(4)
```

```{r functions-square-root-check}
grade_code("Correct! The square root of 4 is 2.")
```

To learn the arguments accepted by a function, call the `args()` function and
pass it the name of the function (without parentheses). Below I've passed
`args()` the function for standard deviation: `sd`.

**Try it: Explore function arguments**
Use the "Run Code" button to see how many arguments this function has and what
they are:

```{r functions-explore-args, exercise=TRUE, exercise.cap="Interactive exercise: Examine the arguments accepted by the sd function using the args function. Use the Run Code button to execute."}
args(sd)
```

This function takes two arguments: `x` and `na.rm`. R matches inputs to
arguments in the order listed. However, you can use the name of the argument
when calling a function, in which case you can list them in any order you like.
Typically R users specify names of arguments for all arguments after the first.
Some arguments have default values and are optional. If the argument is given a
value, as `na.rm` above, you need only include the argument if you wish to
change the default behavior. Other arguments, those without default values, are
required. If you do not specify them, you will get an error message.
We can thus ask for the standard deviation of an object containing the numbers 1
through 5 called **num.list** in three equivalent ways.

**Try it: Different ways to call functions**
Use the "Run Code" button to see how these three function calls produce the same
result:

```{r functions-argument-order, exercise=TRUE, exercise.cap="Interactive exercise: Practice calling functions with arguments in different orders to see equivalent results. Use the Run Code button to execute and compare outputs."}
sd(num.list, FALSE)
sd(na.rm = FALSE, x = num.list)
sd(num.list)
```

Anytime you wish to know about a function -- the arguments it takes, what it
does, the output it returns --  you can use the `help()` function. The help
function takes the name of a function as its argument. Or you can simply type
`?` followed by the name of the function *with no space between ? and the
function name and without parentheses following the function name*. For example,
to learn about the function `sd()` we can do the following (Note: Your browser
may open another window with this information while executing this tutorial):

**Try it: Access function help**
Use the "Run Code" button to get help documentation for the sd function:

```{r functions-get-help, exercise=TRUE, exercise.cap="Interactive exercise: Access help documentation for the sd function using the question mark operator. Use the Run Code button to open help documentation."}
?sd
```

R Studio will open a help window providing details on the function when you run
this function on you local machine.

(If you don't know the name of the R function, google it.)

Note: If you receive an error message like the following, then the function is
likely located in a package you have not loaded: "No documentation for
'funcname' in specified packages and libraries: you could try '??funcname'".
Typing ??funcname will cause R Studio to look more broadly for the function and
indicate which packages contain that function. You can try this option or you
can install the relevant package (if you have not), load it with
library(package_name), and then type ?funcname in the console window.

## Classes of objects

R recognizes different types of objects by assigning each object to a *class*.
This allows R to perform appropriate operations depending on the class of the
object. We've seen two classes of objects so far, numeric and character. In
RStudio, the Environment window will show the class of each object. You can also
use the function `class()`, inserting the object name in parentheses. R will
return the object's class.  Functions, like sqrt and class, have their own class
-- function!

**Practice: Determine object classes**
Find the class of **my.first.object**, **another.object**, and the function `sqrt`. Use one line of code for each object. Enter your code and click "Submit Answer" to check your work.

```{r prepare-classes, echo=FALSE}
my.first.object <- 50 + 30
another.object <- "100"
```

```{r classes-determine-types, exercise=TRUE, exercise.setup = "prepare-classes", exercise.cap="Interactive exercise: Use the class function to determine the type of different R objects. Enter three lines of code and click Submit Answer to check correctness. One hint available."}

```

```{r classes-determine-types-hint}
# Hint: Function Syntax
# You need three separate lines of code. Each line should use the class() function 
# with a different object name inside the parentheses. 
# The three objects are: my.first.object, another.object, and sqrt.
```

```{r classes-determine-types-solution}
# Complete Solution
class(my.first.object)
class(another.object)
class(sqrt)
```

```{r classes-determine-types-check}
grade_code("It's important to always know the class of an object you create.")
```

There are many other classes of objects in R. We will encounter more in future
lessons.

## Vectors

Next, we will enter some data into R as a *numeric vector object*. A *vector* is
just a collection of information stored in a specific order. We use the function
`c()`, which stands for `concatenate,' to enter a data vector containing
multiple values with commas separating different elements of the vector we are
creating. **Any time you wish to refer to more than one element, you must use
the concatenate function!**

**Practice: Create your first vector**
Create a vector called **my.first.vector** that contains any 5 numbers you wish and check the contents of the vector (by listing the name of the vector on a new line) to see your handiwork. Enter your code and click "Submit Answer" to check your work.

```{r vectors-create-first, exercise=TRUE, exercise.cap="Interactive exercise: Create a vector with 5 numbers using the c function and display its contents. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r vectors-create-first-hint}
# Hint: Vector Basics
# Use the c() function to combine multiple numbers, separated by commas.
# Don't forget to assign the result to the specified object name and then display it.
```

```{r vectors-create-first-solution}
# Complete Solution
my.first.vector <- c(1, 2, 3, 4, 5)
my.first.vector
```
Let's try something a bit more interesting.

Here are the two-party vote shares won by the Democratic presidential candidate
in the last 12 presidential elections:

<div class="table-container">
  <table role="table" aria-label="Presidential Election Results from 1980 to 2024 showing Democratic vote shares and candidates">
    <caption>
      <strong>Presidential Election Results 1980-2024</strong><br>
      This table displays the two-party vote share won by Democratic presidential candidates, along with the names of both Democratic and Republican candidates for each election year.
    </caption>
    <thead>
      <tr>
        <th scope="col" id="year">Election Year</th>
        <th scope="col" id="dem-share">Democratic 2-Party Vote Share</th>
        <th scope="col" id="dem-candidate">Democratic Candidate</th>
        <th scope="col" id="rep-candidate">Republican Candidate</th>
      </tr>
    </thead>
    <tbody>
      <tr><th scope="row" headers="year">1980</th><td headers="dem-share">0.448</td><td headers="dem-candidate">Jimmy Carter</td><td headers="rep-candidate">Ronald Reagan</td></tr>
      <tr><th scope="row" headers="year">1984</th><td headers="dem-share">0.408</td><td headers="dem-candidate">Walter Mondale</td><td headers="rep-candidate">Ronald Reagan</td></tr>
      <tr><th scope="row" headers="year">1988</th><td headers="dem-share">0.461</td><td headers="dem-candidate">Michael Dukakis</td><td headers="rep-candidate">George H.W. Bush</td></tr>
      <tr><th scope="row" headers="year">1992</th><td headers="dem-share">0.535</td><td headers="dem-candidate">Bill Clinton</td><td headers="rep-candidate">George H.W. Bush</td></tr>
      <tr><th scope="row" headers="year">1996</th><td headers="dem-share">0.547</td><td headers="dem-candidate">Bill Clinton</td><td headers="rep-candidate">Bob Dole</td></tr>
      <tr><th scope="row" headers="year">2000</th><td headers="dem-share">0.503</td><td headers="dem-candidate">Al Gore</td><td headers="rep-candidate">George W. Bush</td></tr>
      <tr><th scope="row" headers="year">2004</th><td headers="dem-share">0.488</td><td headers="dem-candidate">John Kerry</td><td headers="rep-candidate">George W. Bush</td></tr>
      <tr><th scope="row" headers="year">2008</th><td headers="dem-share">0.537</td><td headers="dem-candidate">Barack Obama</td><td headers="rep-candidate">John McCain</td></tr>
      <tr><th scope="row" headers="year">2012</th><td headers="dem-share">0.520</td><td headers="dem-candidate">Barack Obama</td><td headers="rep-candidate">Mitt Romney</td></tr>
      <tr><th scope="row" headers="year">2016</th><td headers="dem-share">0.511</td><td headers="dem-candidate">Hillary Clinton</td><td headers="rep-candidate">Donald Trump</td></tr>
      <tr><th scope="row" headers="year">2020</th><td headers="dem-share">0.513</td><td headers="dem-candidate">Joe Biden</td><td headers="rep-candidate">Donald Trump</td></tr>
      <!-- New row -->
      <tr><th scope="row" headers="year">2024</th><td headers="dem-share">0.493</td><td headers="dem-candidate">Kamala Harris</td><td headers="rep-candidate">Donald Trump</td></tr>
    </tbody>
  </table>
</div>


We can enter the share of the two-party vote won by the Democratic presidential
candidates in the last 12 elections as elements of a single vector we will name
**dem.pres.vote.share**.

**Practice: Create a data vector**
Create the vector and print it to check your work. Enter your code and click "Submit Answer" to check your work.

```{r vectors-election-data, exercise=TRUE, exercise.cap="Interactive exercise: Create a vector containing Democratic vote shares from the election data table. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r vectors-election-data-hint}
# Hint: Data Entry
# Look at the Democratic 2-Party Vote Share column in the table above.
# Enter the values in chronological order, starting with 1980 and ending with 2024.
# Use the c() function to create a vector with the object name dem.pres.vote.share.
# Remember to print the vector to check your work.
```

```{r vectors-election-data-solution}
# Complete Solution
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488, 0.537, 0.520, 0.511, 0.513, 0.493)
dem.pres.vote.share
```

```{r vectors-election-data-check}
grade_code(" ")
```

The `c()` function can be used to combine multiple vectors.

**Practice: Combine vectors**
Let's create two string vectors, one with the Democratic candidate last names for the first 6 elections (call it **first.6.candidates**) and the other with candidate last names for the next 6 elections (call it **last.6.candidates**) and then combine them into a vector called **dem.candidates**. Check your work by printing the final vector. (Hint: candidates' names are strings. You will need to use quotation marks around each name.) Enter your code and click "Submit Answer" to check your work.

```{r vectors-combine-names, exercise=TRUE, exercise.cap="Interactive exercise: Create and combine two string vectors containing candidate names. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r vectors-combine-names-hint}
# Hint: Vector Combination
# Create two separate vectors first, then combine them.
# Look at the table to see which candidates go in each group (1980-2000 vs 2004-2024).
# Use quotation marks around each name and the c() function to combine the final vectors.
```

```{r vectors-combine-names-solution}
# Complete Solution
first.6.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton", "Clinton", "Gore")
last.6.candidates <- c("Kerry", "Obama", "Obama", "Clinton", "Biden", "Harris")
dem.candidates <- c(first.6.candidates, last.6.candidates)
dem.candidates
```

```{r vectors-combine-names-check}
grade_code("")
```

### Accessing Elements of a Vector

**To select a specific element in a vector, we use square brackets [ ].** We insert the *index* associated with the element of the vector we want to select. The *index* value 1 refers to the first element in the vector. The *index* value 2 refers to the second element in the vector, and so on. You can extract any element of a vector by placing a pair of brackets [ ] after the vector name. Inside the brackets, place the number of the element that you'd like to extract. Here is an example that returns the first value in **dem.pres.vote.share**. Change it to extract Hillary Clinton's vote share.

**Practice: Access a single vector element**
Modify this code to extract Hillary Clinton's vote share (the 10th element). Enter your code and click "Submit Answer" to check your work.

```{r vectors-access-single, exercise=TRUE, exercise.cap="Interactive exercise: Extract a specific element from a vector using index notation. Enter your code and click Submit Answer to check correctness. One hint available."}
dem.pres.vote.share[1]
```

```{r vectors-access-single-hint}
# Hint: Index Position
# Hillary Clinton's data is the 10th element in the vector. 
# Change the number inside the square brackets from 1 to 10.
```

```{r vectors-access-single-solution}
# Complete Solution
dem.pres.vote.share[10]
```

```{r vectors-access-single-check}
grade_code(" ")
```

We can extract more than one element by using the `c()` function inside of the square brackets.

**Practice: Access multiple vector elements**
What was the vote share won by the Democrat in the last 5 elections? Enter your code and click "Submit Answer" to check your work.

```{r vectors-access-multiple, exercise=TRUE, exercise.setup="vectors-election-data", exercise.cap="Interactive exercise: Extract multiple elements from a vector using the c function within brackets. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r vectors-access-multiple-hint}
# Hint: Recent Elections
# The last 5 elections span from 2008 to 2024.
# Count backwards from the end of the vector to find the correct index positions.
```

```{r vectors-access-multiple-solution}
# Complete Solution
dem.pres.vote.share[c(8, 9, 10, 11, 12)]
```

```{r vectors-access-multiple-check}
grade_code(" ")
```

Note that the order in which we list the indexes matters.

**Try it: Order of vector elements**
Use the "Run Code" button to see how the order of index numbers affects the
output:

```{r vectors-order-matters, exercise=TRUE, exercise.setup="vectors-election-data", exercise.cap="Interactive exercise: Observe how the order of index values affects vector output. Use the Run Code button to execute and compare the different results."}
dem.pres.vote.share[c(9, 5)]
dem.pres.vote.share[c(5, 9)]
```

We can also use a dash in the index to remove an element from a vector. For
example, we can remove the vote share won by the Democrat in the last election (2024, which is the 12th element) as follows:

**Try it: Remove vector elements**
Use the "Run Code" button to see how to exclude specific elements from a vector
using negative indexing:

```{r vectors-remove-elements, exercise=TRUE, exercise.cap="Interactive exercise: Remove specific elements from a vector using negative indexing. Use the Run Code button to execute and see the result."}
dem.pres.vote.share[-12]
```

## Numeric Operations on a Numeric Vector

We can apply arithmetic operations on a numeric vector. *The operation will be repeated for each element of the vector.* We will often want to do this as part of creating new variables that are a function of variables in a data set.

**Practice: Convert vote shares to percentages**
See if you can create a new vector called **dem.pres.vote.percent** that converts our vote *share* vector (**dem.pres.vote.share**) to the *percent* of the 2-party vote won by the Democratic candidate and print the output. Enter your code and click "Submit Answer" to check your work.

```{r vectors-convert-percent, exercise=TRUE, exercise.setup="vectors-election-data", exercise.cap="Interactive exercise: Convert vote shares to percentages by multiplying by 100. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r vectors-convert-percent-hint}
# Hint: Unit Conversion
# To convert from decimal shares to percentages, multiply by 100.
# Assign the result to a new vector and remember to display it.
```

```{r vectors-convert-percent-solution}
# Complete Solution
dem.pres.vote.percent <- dem.pres.vote.share * 100
dem.pres.vote.percent
```

```{r vectors-convert-percent-check}
grade_code(" ")
```

We can also create a vector that expresses the Democratic candidate's vote share compared to that in 1980. In other words, we can subtract the 1980 vote share value from each element in the vector of democratic vote shares. See if you can figure out how to do this without looking at the hint. Name your new vector **compared.to.1980** and print the results.

**Practice: Calculate differences from baseline**
Create a vector showing how each election's Democratic vote share compares to 1980. Enter your code and click "Submit Answer" to check your work.

```{r vectors-compare-baseline, exercise=TRUE, exercise.setup="vectors-election-data", exercise.cap="Interactive exercise: Calculate the difference between each election and the 1980 baseline. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r vectors-compare-baseline-hint}
# Hint: Baseline Comparison
# Use vector indexing to access the 1980 value (first element).
# Subtract this baseline from the entire vector to see differences.
```

```{r vectors-compare-baseline-solution}
# Complete Solution
compared.to.1980 <- dem.pres.vote.share - dem.pres.vote.share[1]
compared.to.1980
```

```{r vectors-compare-baseline-check}
grade_code()
```

Note that positive numbers tell us the current candidate did better than the Democratic candidate in 1980, negative numbers tell us that candidate did worse.

## Replacing Elements in a Vector

We can also replace values associated with a particular index using the assignment operator. Let's say we want to replace the three "Clinton"s with first and last names to clarify the name of the Democratic candidate.

**Practice: Replace vector elements**
Can you figure out how to replace the three "Clinton" entries with their full names? (Remember to check your work by printing your object.) Enter your code and click "Submit Answer" to check your work.

```{r vectors-replace-elements, exercise=TRUE, exercise.setup="vectors-combine-names", exercise.cap="Interactive exercise: Replace specific vector elements using index assignment. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r vectors-replace-elements-hint}
# Hint: Element Replacement
# Look at the dem.candidates vector to find where "Clinton" appears.
# Use index positions with square brackets to replace multiple elements at once.
# The three Clintons are at positions 4, 5, and 10.
```

```{r vectors-replace-elements-solution}
# Complete Solution
dem.candidates[c(4,5,10)] <- c("Bill Clinton", "Bill Clinton", "Hillary Clinton")
dem.candidates
```

```{r vectors-replace-elements-check}
grade_code(" ")
```

## Back to functions

Some functions you will use a lot include:

* `length()`: tells us the length of the vector or number of elements it has.
* `min()`: tells us the minimum value.
* `max()`: returns the maximum value.
* `range()`: returns the range of the data.
* `mean()`: provides the mean of the object values.
* `sum()`: provides the sum of the values in the object.

Each of these functions has two arguments: `x`, the name of the object for which
the function is to be applied and `na.rm`, which stands for remove missing
values and takes a default value of FALSE. If the object contains missing
values, these functions will return an error unless you specify `na.rm = TRUE`.
This tells R to ignore the missing values before applying the function. None of
the vectors we created have missing values, so the default option is fine.

**Try it: Apply summary functions**
Below I've applied each function to some (numeric!) vector we created with our
vote data. Use the "Run Code" button to execute the functions and see the
results:

```{r functions-summary-stats, exercise=TRUE, exercise.cap="Interactive exercise: Apply various summary functions to vector data. Use the Run Code button to execute and observe the results of different statistical functions."}
length(dem.candidates)
min(dem.pres.vote.share)
max(dem.pres.vote.share)
range(dem.pres.vote.share)
mean(dem.pres.vote.share)
sum(dem.pres.vote.share) / length(dem.pres.vote.share)
```

The last expression gives another way of calculating the mean as the sum of all
the elements divided by the number of elements.

### Functions with Multiple Inputs

If a function requires multiple inputs, the syntax looks like `funcname(input1,
input2)`. The order of the input matters. To avoid confusion and problems
stemming from getting the order wrong, it is a good idea to specify the name of
the argument that each input corresponds to (you can find a list of arguments by
using the help function -- or in most cases, RStudio will show this information
as you type the name of the function). This looks like `funcname(arg1=input1,
arg2=input2)`.

As an example, the `seq()` function can generate a vector composed of an
increasing or decreasing sequence of numbers. The first argument `from`
specifies the number to start from; the second argument `to` specifies the
number at which to end the sequence; the last argument `by` indicates the
interval to increase or decrease by.

**Try it: Create number sequences**
Here's an example that creates a sequence of numbers from 1 to 10 by increments
of 2 and checks our work. Use the "Run Code" button to execute:

```{r functions-create-sequence, exercise=TRUE, exercise.cap="Interactive exercise: Create a numeric sequence using the seq function with specified intervals. Use the Run Code button to execute and see the sequence generated."}
my.sequence <- seq(from = 1, to = 10, by = 2)
my.sequence
```

Let's create an object for the "election year" variable in our table using this
function.

**Practice: Create election year sequence**
Name your object **elect.year** and print the result. Enter your code and click "Submit Answer" to check your work.

```{r functions-election-years, exercise=TRUE, exercise.cap="Interactive exercise: Create a sequence of election years from 1980 to 2024. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r functions-election-years-hint}
# Hint: Sequential Years
# Election years run from 1980 to 2024, with elections every 4 years.
# Use the seq() function with appropriate from, to, and by arguments.
```

```{r functions-election-years-solution}
# Complete Solution
elect.year <- seq(from = 1980, to = 2024, by = 4)
elect.year
```

```{r functions-election-years-check}
grade_code(" ")
```

Note that we can create decreasing sequences by making the `by` argument negative.

**Try it: Create decreasing sequences**
Use the "Run Code" button to see how to create a sequence that counts backwards:

```{r functions-decreasing-sequence, exercise=TRUE, exercise.cap="Interactive exercise: Create a decreasing numeric sequence using negative by values. Use the Run Code button to execute and see the decreasing sequence."}
seq(from = 2024, to = 1980, by = -4)
```

The `names()` function can access and assign names to an object. Names are not part of the data themselves but are a useful attribute of the R object.

Below we use the `names()` function first to see if names are assigned to the vector **dem.pres.vote.share** and seeing that there are none, we assign the names of the election year to the **dem.pres.vote.share.** Once we do so, the vector is printed with a useful label.

```{r functions-prepare-names, echo=FALSE}
elect.year <- seq(from = 1980, to = 2024, by = 4)
```

**Try it: Assign and use names**
Use the "Run Code" button to see how to add descriptive names to vector
elements:

```{r functions-assign-names, exercise=TRUE, exercise.setup="functions-prepare-names", exercise.cap="Interactive exercise: Add descriptive names to vector elements using the names function. Use the Run Code button to execute and see how names are assigned and displayed."}
names(dem.pres.vote.share)
names(dem.pres.vote.share) <- elect.year
names(dem.pres.vote.share)
dem.pres.vote.share
```

## Packages

Base R contains many functions to accomplish a variety of tasks. But many of the
functions we will use are contained in packages that have been contributed to
the Comprehensive R Archive Network (CRAN).  A package bundles together code,
data, and help files. There are over 10,000 packages written for R, and more are
contributed all the time. R does not automatically install or load packages in
order to save memory.

You will need to install packages using the `install.packages()` function by
passing the function the name of the package in quotes. Once installed, you load
the package using the `library()` function. For example, we will use the
`ggplot()` function in the  <span style="color:DarkGreen">ggplot2</span>
package. You must load a package each time you start R before you can use the
functions it contains (but only install it once). If you don't first load the
package, you will get an error message:

**Try it: See package error**
Use the "Run Code" button to observe what happens when you try to use a function
from an unloaded package:

```{r packages-error-demo, exercise=TRUE, exercise.cap="Interactive exercise: Observe error message when trying to use function from unloaded package. Use the Run Code button to execute and see the error."}
ggplot()
```

Instead, we load the package first. This makes its functions available in
memory. Note the error message. Any time you receive an error that begins "could
not find function" the likely problem is that you haven't loaded the packaged
that contains the function!

**Try it: Load package and use function**
Now let's see what happens when we properly load the package first. You'll see a blank gray plot area when you run this code - that's normal! The `ggplot()` function creates an empty plotting canvas waiting for data and plot specifications. The key difference from the previous exercise is that the function runs without error, proving the package loaded successfully.

```{r packages-load-demo, exercise=TRUE, exercise.cap="Interactive exercise: Load a package using library and then access its functions. Use the Run Code button to execute and see successful function access."}
library(ggplot2)
ggplot()
```

To get help with a package type the `help()` function and pass it the `package`
argument. Set the argument equal to the name of the package in quotes. You don't
need to read the output below):

**Try it: Get package help**
Use the "Run Code" button to learn how to access comprehensive help for an
entire package:

```{r packages-get-help, exercise=TRUE, exercise.cap="Interactive exercise: Access comprehensive help documentation for an R package. Use the Run Code button to execute and open package documentation."}
help(package = "ggplot2")
```

## Comments in code

If you wish to add comments to your code to help you remember why you ran a
function or the arguments it takes, you place a hashtag before the comment.
Anything after # on the same line will not be run.

**Try it: Use code comments**
Use the "Run Code" button to see how comments can document your code without
affecting execution:

```{r comments-demo, exercise=TRUE, exercise.cap="Interactive exercise: Add explanatory comments to R code using hashtag symbols. Use the Run Code button to execute and see how comments are ignored."}
#Sample comment
mean(dem.pres.vote.share, na.rm = TRUE)
```

## Practice

The following exercises combine concepts you've learned throughout this tutorial. Take your time and think through each problem step by step.

**Practice: Calculate Republican vote shares**
Create a vector called **rep.pres.vote.share** containing the share of the two-party vote received by the Republican presidential candidate. Print it. Note that you do NOT need to type in the vote shares because we are only considering votes for the major parties. That means the Republican vote share is just one minus the Democratic vote share. Enter your code and click "Submit Answer" to check your work.

```{r practice-republican-shares, exercise=TRUE, exercise.cap="Interactive exercise: Calculate Republican vote shares as one minus Democratic shares. Enter your code and click Submit Answer to check correctness. Two hints available."}

```

```{r practice-republican-shares-hint-1}
# Hint 1 of 2: Two-Party Math
# In a two-party system, the vote shares must add up to 1 (or 100%).
# If Democrats got some share, Republicans got the remainder.
```

```{r practice-republican-shares-hint-2}
# Hint 2 of 2: Implementation
# Use arithmetic: 1 minus the Democratic vote share vector.
# Assign the result to rep.pres.vote.share and print it.
```

```{r practice-republican-shares-solution}
# Complete Solution
rep.pres.vote.share <- 1 - dem.pres.vote.share
rep.pres.vote.share
```

```{r practice-republican-shares-check}
grade_code(" ")
```

**Practice: Calculate victory margins**
Create a vector called **victory.margin** that calculates the margin of victory. Here you will need to use the absolute value function, which is the `abs()` function. (You will need to list the Republican vote share first inside this function to match my answer, but you could do either in your own work.) Note that we don't know which party won, we simply know the proportion of the two-party vote by which the winner won. Print it. Enter your code and click "Submit Answer" to check your work.

```{r practice-setup-for-margins, echo=FALSE}
rep.pres.vote.share <- 1 - dem.pres.vote.share
```

```{r practice-victory-margins, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Interactive exercise: Calculate victory margins using absolute differences between party vote shares. Enter your code and click Submit Answer to check correctness. Two hints available."}

```

```{r practice-victory-margins-hint-1}
# Hint 1 of 2: Victory Margin Concept
# Victory margin is how much the winner beat the loser by.
# It's the absolute difference between the two vote shares.
```

```{r practice-victory-margins-hint-2}
# Hint 2 of 2: Absolute Difference
# Use abs() around the subtraction of the two vote share vectors.
# Assign to victory.margin and print the result.
```

```{r practice-victory-margins-solution}
# Complete Solution
victory.margin <- abs(rep.pres.vote.share - dem.pres.vote.share)
victory.margin
```

```{r practice-victory-margins-check}
grade_code(" ")
```



**Practice: Create Republican candidate names**
Create a vector of Republican candidate names called **rep.names.** Print it. Just use last names. Enter your code and click "Submit Answer" to check your work.

```{r practice-republican-names, exercise=TRUE, exercise.cap="Interactive exercise: Create a vector containing Republican presidential candidate last names. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r practice-republican-names-hint}
# Hint: Candidate Names
# Use the c() function with Republican candidate last names in quotes.
# Check the election table earlier in the tutorial for the chronological list from 1980-2024.
```

```{r practice-republican-names-solution}
# Complete Solution
rep.names <- c("Reagan", "Reagan", "Bush", "Bush", "Dole", "Bush", "Bush", "McCain", "Romney", "Trump", "Trump", "Trump")
rep.names
```

```{r practice-republican-names-check}
grade_code(" ")
```

**Practice: Calculate mean Republican vote share**
Find the mean share of the vote going to the Republican candidate (there are no missing values, so you do not need to specify the `na.rm` argument). Enter your code and click "Submit Answer" to check your work.

```{r practice-mean-republican, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Interactive exercise: Calculate the mean Republican vote share across all elections. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r practice-mean-republican-hint}
# Hint: Average Calculation
# Use the function that calculates the average of a set of numbers.
# Apply it to the Republican vote share vector.
```

```{r practice-mean-republican-solution}
# Complete Solution
mean(rep.pres.vote.share)
```

```{r practice-mean-republican-check}
grade_code(" ")
```

**Practice: Find Republican vote share range**
Find the range of the vote share received by Republican candidates. Enter your code and click "Submit Answer" to check your work.

```{r practice-range-republican, exercise=TRUE, exercise.setup="practice-setup-for-margins", exercise.cap="Interactive exercise: Find the minimum and maximum Republican vote shares using the range function. Enter your code and click Submit Answer to check correctness. One hint available."}

```

```{r practice-range-republican-hint}
# Hint: Min and Max
# Use the function that shows both the minimum and maximum values.
# Apply it to the Republican vote share vector.
```

```{r practice-range-republican-solution}
# Complete Solution
range(rep.pres.vote.share)
```

```{r practice-range-republican-check}
grade_code(" ")
```

## The Takeaways

This tutorial introduced you to the fundamental building blocks of working with data in R. These concepts form the foundation for all data analysis work you'll do throughout the semester. We've learned:

1. **R can serve multiple roles in data analysis**: R functions as both a powerful calculator for mathematical operations and a programming environment for storing, manipulating, and analyzing data. Understanding this dual nature is essential for effective data work.

2. **Objects are the foundation of R programming**: Everything in R is stored as objects with specific names and classes. Key principles include:

   a. **Object naming**: Choose informative names that cannot start with numbers, contain spaces, or include special characters like $ or #.
   b. **Assignment**: Use the `<-` operator to assign values to objects.
   c. **Object classes**: R recognizes different types of data (numeric, character, function) and treats them appropriately. Always verify object classes using `class()`.

3. **Vectors are essential data structures**: Most data analysis involves working with collections of related information stored as vectors. Critical vector concepts include:

   a. **Creation**: Use `c()` to combine multiple elements into a single vector.
   b. **Indexing**: Access specific elements using square brackets `[]` with index numbers starting from 1.
   c. **Operations**: Arithmetic operations applied to vectors affect each element, enabling efficient data transformations.

4. **Functions extend R's capabilities**: Functions take inputs (arguments) and return outputs, allowing you to perform complex operations efficiently. Essential function concepts include:

   a. **Syntax**: Functions follow the pattern `function_name(arguments)` with commas separating multiple arguments.
   b. **Help**: Use `?function_name` or `help()` to learn about any function's arguments and behavior.
   c. **Packages**: Install packages with `install.packages()` and load them with `library()` to access additional functions beyond base R.

5. **Attention to detail is crucial**: R requires precise syntax including matching parentheses, correct spelling, proper comma placement, and consistent object names. Most errors stem from these mechanical issues rather than conceptual misunderstandings.

The skills you've practiced here—creating objects, working with vectors, using functions, and understanding R's syntax requirements—will be essential in every subsequent tutorial and assignment. Master these fundamentals, as they form the building blocks for more complex data analysis techniques throughout the course.

<div class="success-note"> <strong>Congratulations!</strong> Way to go! You've
finished the first tutorial. </div>