---
title: "R Basics: Part 1"
tutorial:
  id: "01-R-Part1"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn the basics of how R works.
--- 

## Introduction

This is the first of a series of tutorials you will work through over the course of the semester.  This tutorial is designed to familiarize you with some of the basic principles of using the statistical software package R and provide background knowledge that will underlie the code, the instructions the software uses to extract some information from the data, we write later in the semester.

The tutorials will typically introduce code with examples that you will run -- using the "Run Code" button -- or submit your code to be checked -- using the "Submit Answer" button.  Then you will be asked to edit the code or complete partial code to perform related tasks and run that code.  Finally, you will be asked to reproduce code from scratch.  A "Hint" or "Solution" button will appear to help you if you are stuck, as well as the "Run Code" button.  Don't be afraid to try out code before you look at hints.

<span style="color:blue">You will make mistakes. R, like any coding language, requires meticulous attention to things like matching parentheses, including commas in the right places, and spelling functions, arguments, and object names correctly. When you encounter errors -- and you will do so often -- always look for these types of mistakes first!>/span>

You should work through the tutorials beginning to end, completing each exercise. If you run out of time, you can exit and return to the tutorial later but unless you are running the tutorials on a local installation of R Studio, your progress will not be saved. 

I encourage you to resist the temptation to view the solution and copy it. Typing out the code will help you to understand what each exercise is doing and to recall code in you own work. Take notes as you work through the tutorials. A good strategy would be to create a list of the functions introduced, note the name of the tutorial that introduced them, and write a short description. This will provide a help sheet to reference when you need to write your own code from scratch in class, for labs, or for your final project.


## Learning Objectives

This tutorial is designed to familiarize you with some of the basics of using R. Specifically, you will learn:

* How to use R as a calculator
* How to assign a value or values to an object with <-
* How to use functions
* How to determine the class of an object with the `class()` function
* How to create and use vectors using the `c()` function
* How to select and replace elements of a vector using indexing
* How to install packages with the `install.packages()` function and access the functions in them with the `library()` function



```{r setup, include=FALSE}
library(learnr)
#library(tidyverse)
library(knitr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
#tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488, 0.537, 0.520, 0.511, 0.513)
first.5.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton","Clinton")
last.6.candidates <- c("Gore", "Kerry","Obama","Obama", "Clinton", "Biden")
dem.candidates <- c(first.5.candidates, last.6.candidates)
num.list<-c(1,2,3,4,5)
```


## R as a calculator

It is easy to use R like a calculator. Here's an example of adding 2 numbers. You can add/subtract, multiply/divide, or raise a number to a power using `+`, `-`, `*`, `/`, `^`. Run the code and then change and rerun it to practice using R as a calculator.  Note that R ignores spaces, so 14 + 2000 will produce the same output as 14+2000.



```{r calc, exercise=TRUE}
14+2000
```


The `[1]` you see with the line of output is just R telling us that we have just one number produced by the code. Later we will see cases where a command we run produces an output that contains multiple numbers (or words).  Each line of output will begin with some number in [ ], with the number denoting the index value of the output that starts that line.



## Objects


**R can store information as an object with a name of your choice. Once you have created an R object, you can refer to it by name.**

Picking a name for your object:

* Choose an informative name
* Object names cannot start with a number
* Object names cannot contain spaces
* Object names cannot include special characters that have specific meanings in R. This includes $ and # and %.


The Global Environment window in R Studio will show a list of all the objects you create.

**We use the assignment operator `<-` to assign some value to an object.**

Let's assign the sum of 5+3 to an object called **my.first.object**. Run the following code.

```{r firstobj, exercise=TRUE}
my.first.object <- 5+3
```

Notice that the result is not displayed in this case. We can see the value of an object only by asking R to show it to us. Simply enter the object name.

```{r printfirstobj, exercise=TRUE}
my.first.object <- 5+3
my.first.object
```


If we assign a different value to the same object name, the value of the object will change.

```{r replace, exercise=TRUE}
my.first.object <-50+30
my.first.object
```

Object names are case sensitive: my.first.object is different from **MY.FIRST.OBJECT**. Try asking R to tell you the value of **MY.FIRST.OBJECT**. What happens?

```{r error1, exercise=TRUE}
MY.FIRST.OBJECT
```


You will see error messages a lot! It's part of the learning process. The trick is to learn how to fix them! In this case, the error message is pretty clear. Looking in the Global Environment in RStudio you would see that **MY.FIRST.OBJECT** does not exist. 


###  Storing strings (characters/words etc.) as objects

To assign a string of characters to an object, we use quotation marks. Spaces ARE allowed.

```{r string, exercise=TRUE}
this.class <- "PLSC 309"
this.class
```

R treats numbers like characters if we tell it to do so by enclosing them in quotes:

```{r string2, exercise=TRUE}
another.object <- "100"
another.object
```

If you refer to an object that is a character string in an arithmetic operation, you will see an error message. See what happens if you try to divide "100" by 10:

```{r error2, exercise=TRUE}
another.object <- "100"
another.object /  10
```



## Functions

Functions are very important in R. They can be used to perform a variety of tasks. **A function takes one or more inputs and returns some output.**

For example, the `sqrt()` function takes as its input some number or numeric object for which we want to know the square root. The output is the square root of that number or object. In R, the syntax is generally `funcname(input)`, where `funcname` is the function name and `input` is one or more *arguments*. Some functions take no arguments, some take one argument, and others take several arguments. When there is more than one argument, they are separated by a comma. Note that omitting a required comma and having unbalanced parentheses are frequent sources of error.  

The function `sqrt()` takes one argument, named `x`, which is the number or object for which we want the square root. Try taking the square root of 4:

```{r squareroot, exercise=TRUE}

```

```{r squareroot-hint}
Place the number 4 inside the sqrt function
```

```{r squareroot-solution}
sqrt(4)
```

```{r squareroot-check}
grade_code("Not so hard.")
```

To learn the arguments accepted by a function, call the `args()` function and pass it the name of the function (without parentheses). Below I've passed `args()` the function for standard deviation: `sd`. Run the code below to see how many arguments this function has and what they are.

```{r args, exercise=TRUE}
args(sd)
```

This function takes two arguments: `x` and `na.rm`. R matches inputs to arguments in the order listed. However, you can use the name of the argument when calling a function, in which case you can list them in any order you like. Typically R users specify names of arguments for all arguments after the first. 

Some arguments have default values and are optional. If the argument is given a value, as `na.rm` above, you need only include the argument if you wish to change the default behavior. Other arguments, those without default values, are required. If you do not specify them, you will get an error message.

We can thus ask for the standard deviation of an object containing the numbers 1 through 5 called **num.list** in three equivalent ways.

```{r order, exercise=TRUE}
sd(num.list, FALSE)
sd(na.rm=FALSE, x=num.list)
sd(num.list)
```



Anytime you wish to know about a function -- the arguments it takes, what it does, the output it returns --  you can use the `help()` function. The help function takes the name of a function as its argument. Or you can simply type `?` followed by the name of the function *without parentheses*. For example, to learn about the function `sd()` we can do the following (Note: Your browser may open another window with this information while executing this tutorial):

```{r helper, exercise=TRUE}
?sd
```

R Studio will open a help window providing details on the function.

(If you don't know the name of the R function, google it.) 




## Classes of objects

R recognizes different types of objects by assigning each object to a *class*. This allows R to perform appropriate operations depending on the class of the object. We've seen two classes of objects so far, numeric and character. In RStudio, the Environment window will show the class of each object. You can also use the function `class()`, inserting the object name in parentheses. R will return the object's class.  Functions, like sqrt and class, have their own class -- function!


Find the class of **my.first.object**, **another.object**, and the function `sqrt`:


```{r prepare-classes, echo=FALSE}
my.first.object <-50+30
another.object <- "100"
```

```{r classes, exercise=TRUE,  exercise.setup = "prepare-classes"}

```

```{r classes-hint}
This will require three lines of code. In each type class and insert
the name of the object in parentheses.
```

```{r classes-solution}
class(my.first.object)
class(another.object)
class(sqrt)
```

```{r classes-check}
grade_code("It's important to always know the class of an object you create.")
```

There are many other classes of objects in R. We will encounter more in future lessons.


## Vectors

Next, we will enter some data into R as a *numeric vector object*. A *vector* is just a collection of information stored in a specific order. We use the function `c()`, which stands for `concatenate,' to enter a data vector containing multiple values with commas separating different elements of the vector we are creating. **Any time you wish to refer to more than one element, you must use the concatenate function!**

Create a vector called **my.first.vector** that contains any 5 numbers you wish and check the contents of the vector to see your handiwork:


```{r firstvec, exercise=TRUE}

```

```{r firstvec-solution}
my.first.vector <-c(1,2,3,4,5)
my.first.vector
```



Let's try something a bit more interesting.

Here are the two-party vote shares won by the Democratic presidential candidate in the last 11 presidential elections:


Election Year  | Democratic 2-Party Vote Share | Democratic Candidate | Republican Candidate
-------------  | ------------------ | -------------------- | --------------
 1980   | 0.448 | Jimmy Carter | Ronald Reagan
 1984   | 0.408 | Walter Mondale | Ronald Reagan
 1988   | 0.461 | Michael Dukakis | George H.W. Bush
 1992   | 0.535 | Bill Clinton | George H.W. Bush
 1996   | 0.547 | Bill Clinton | Bob Dole
 2000   | 0.503 | Al Gore | George W. Bush
 2004   | 0.488 | John Kerry | George W. Bush
 2008   | 0.537 | Barack Obama | John McCain
 2012   | 0.520 | Barack Obama | Mitt Romney
 2016   | 0.511 | Hillary Clinton | Donald Trump
 2020   | 0.513 | Joe Biden | Donald Trump
 
We can enter the share of the two-party vote won by the Democratic presidential candidates in the last 11 elections as elements of a single vector we will name **dem.pres.vote.share**. 

Create the vector and print it to check your work.

```{r demshare, exercise=TRUE}

```

```{r demshare-hint}
1. Type the name of the object you wish to create and then
type the assignment operator.
2. You will place the values, separated by commas, inside the c() function.
3. Finally, type the name of the object to view the result.
```

```{r demshare-solution}
dem.pres.vote.share <- c(0.448, 0.408, 0.461, 0.535, 0.547, 0.503, 0.488, 0.537, 0.520, 0.511, 0.513)
dem.pres.vote.share
```

```{r demshare-check}
grade_code(" ")
```

The `c()` function can be used to combine multiple vectors. 

Let's create two string vectors, one with the Democratic candidate last names for the first 5 elections (call it **first.5.candidates**) and the other with candidate last names for the second 6 elections (call it **last.6.candidates**) and then combine them into a vector called **dem.candidates**. Check your work by printing the final vector. (Hint: candidates' names are strings. You will need to use quotation marks around each name.)

```{r candnames, exercise=TRUE}

```


```{r candnames-hint}
To create first.5.candidates give the name and then assign it
to the object with <-, then type c() with the first five
names in parentheses, separated by commas. Repeat with the
second six names and then use the c() function to combine 
the two objects and assign the result to dem.candidates.
```

```{r candnames-solution}
first.5.candidates <- c("Carter", "Mondale", "Dukakis", "Clinton","Clinton")
last.6.candidates <- c("Gore", "Kerry","Obama","Obama", "Clinton", "Biden")
dem.candidates <- c(first.5.candidates, last.6.candidates)
dem.candidates
```

```{r candnames-check}
grade_code("")
```


### Accessing Elements of a Vector

**To select a specific element in a vector, we use square brackets [ ].** We insert the *index* associated with the element of the vector we want to select. The *index* value 1 refers to the first element in the vector. The *index* value 2 refers to the second element in the vector, and so on. You can extract any element of a vector by placing a pair of brackets [ ] behind the vector. Inside the brackets, place the number of the element that you’d like to extract. Here is an example that returns the first value in **dem.pres.vote.share** Change it to extract Hillary Clinton's vote share.

```{r access, exercise=TRUE}
dem.pres.vote.share[1]
```

```{r access-hint}
This information is contained in the 10th element in our vector dem.pres.vote.share
```

```{r access-solution}
dem.pres.vote.share[10]
```


```{r access-check}
grade_code(" ")
```


We can extract more than one element by using the `c()` function inside of the square brackets. 

What was the vote share won by the Democrat in the last 5 elections? 

```{r last5, exercise=TRUE, exercise.setup="demshare"}

```

```{r last5-hint}
Inside square brackets use the c() function to
select elections 7 through 11, separating each
by commas
```

```{r last5-solution}
dem.pres.vote.share[c(7,8,9,10,11)]
```

```{r last5-check}
grade_code(" ")
```

Note that the order in which we list the indexes matters.

```{r order2, exercise=TRUE, exercise.setup="demshare"}
dem.pres.vote.share[c(9,5)]
dem.pres.vote.share[c(5,9)]
```

We can also use a dash in the index to remove an element from a vector. For example, we can remove the vote share won by the Democrat in the last election as follows:

```{r remove, exercise=TRUE}
dem.pres.vote.share[-11]
```

## Numeric Operations on a Numeric Vector

We can apply arithmetic operations on a numeric vector. *The operation will be repeated for each element of the vector.* We will often want to do this as part of creating new variables that are a function of variables in a data set.

See if you can create a new vector called **dem.pres.vote.percent** that converts our vote *share* vector (**dem.pres.vote.share**) to the *percent* of the 2-party vote won by the Democratic candidate and print the output:

```{r percent, exercise=TRUE, exercise.setup="demshare"}

```

```{r percent-hint}
Multiply dem.pres.vote.share by 100 and assign the
result to dem.pres.vote.percent

```

```{r percent-solution}
dem.pres.vote.percent <- dem.pres.vote.share * 100
dem.pres.vote.percent
```

```{r percent-check}
grade_code(" ")
```

We can also create a vector that expresses the Democratic candidate's vote share compared to that in 1980. In other words, we can subtract the 1980 vote share value from each element in the vector of democratic vote shares. See if you can figure out how to do this without looking at the hint. Name your new vector **compared.to.1980** and print the results.

```{r subtract, exercise=TRUE, exercise.setup="demshare"}

```

```{r subtract-hint}
We want to subtract the 1980 vote share from each element of the vector dem.pres.vote.share.  The 1980 value is the 1st element of the vector. 
Thus we can substract dem.pres.vote.share[1] from the original vector.
```

```{r subtract-solution}
compared.to.1980 <- dem.pres.vote.share - dem.pres.vote.share[1]
compared.to.1980
```

```{r subtract-check}
grade_code()
```

Note that positive numbers tell us the current candidate did better than the Democratic candidate in 1980, negative numbers tell us that candidate did worse. 

## Replacing Elements in a Vector

We can also replace values associated with a particular index using the assignment operator. Let's say we want to replace the three "Clinton"s with first and last names to clarify the name of the Democratic candidate.  
Can you figure out how to do that without looking at the hints? Try it! (Remember to check your work by printing your object)

```{r rename, exercise=TRUE, exercise.setup="candnames"}

```

```{r rename-hint-1}
First, you need to figure out which elements (which
index values) of your vector of candidate names
(**dem.candidates**) are associated with elections in
which one of the Clintons ran for president. 
```

```{r rename-hint-2}
Use square brackets to surround the index values you just identified. 
Inside those square brackets, you will need to use the concatenate 
function. This specifies the index values you wish to replace.

Then assign the replacement candidate. Be sure to
use the `c()` function, place the names in quotes, and
separate the names with commas.

That's a lot to remember, but you will get the hang of
it, I promise!
```

```{r rename-solution}
dem.candidates[c(4,5,10)] <- c("Bill Clinton", "Bill Clinton", "Hillary Clinton")
dem.candidates
```

```{r rename-check}
grade_code(" ")
```

## Back to functions

Some functions you will use a lot include:

* `length()`: tells us the length of the vector or number of elements it has.
* `min()`: tells us the minimum value.
* `max()`: returns the maximum value.
* `range()`: returns the range of the data.
* `mean()`: provides the mean of the object values.
* `sum()`: provides the sum of the values in the object.

Each of these functions takes only one input object (argument). 

Below I've applied each function to some (numeric!) vector we created with our vote data:

```{r try1, exercise=TRUE}
length(dem.candidates)
min(dem.pres.vote.share)
max(dem.pres.vote.share)
range(dem.pres.vote.share)
mean(dem.pres.vote.share)
sum(dem.pres.vote.share) / length(dem.pres.vote.share)
```

The last expression gives another way of calculating the mean as the sum of all the elements divided by the number of elements.

### Functions with Multiple Inputs

If a function requires multiple inputs, the syntax looks like `funcname(input1, input2)`. The order of the input matters. To avoid confusion and problems stemming from getting the order wrong, it is a good idea to specify the name of the argument that each input corresponds to (you can find a list of arguments by using the help function -- or in most cases, RStudio will show this information as you type the name of the function). This looks like `funcname(arg1=input1, arg2=input2)`.

As an example, the `seq()` function can generate a vector composed of an increasing or decreasing sequence of numbers. The first argument `from` specifies the number to start from; the second argument `to` specifies the number at which to end the sequence; the last argument `by` indicates the interval to increase or decrease by. 

Here's an example that creates a sequence of numbers from 1 to 10 by  increments of 2 and checks our work:

```{r exseq, exercise=TRUE}
my.sequence <- seq(from=1,to=10,by=2)
my.sequence
```


Let's create an object for the "election year" variable in our table using this function. Name your object **elect.year**.

```{r years, exercise=TRUE}

```

```{r years-hint}
The first election year is 1980 and the last is 2020. 
We only have election years to need to increment by 4.
Specity the from, to, and by arguments inside the seq()
function. Don't forget to separate the arguments with commas.
```


```{r years-solution}
elect.year <- seq(from=1980, to=2020, by=4)
elect.year
```

```{r years-check}
grade_code(" ")
```



Note that we can create decreasing sequences by making the `by` argument negative. 

```{r backward, exercise=TRUE}
seq(from=2020, to=1980,
    by=-4)
```


The `names()` function can access and assign names to an object. Names are not part of the data themselves but are a useful attribute of the R object.

Below we use the `names()` function first to see if names are assigned to the vector **dem.pres.vote.share** and seeing that there are none, we assign the names of the election year to the **dem.pres.vote.share.** Once we do so, the vector is printed with a useful label.

```{r elect}
elect.year <- seq(from=1980, to=2020, by=4)
```

```{r names, exercise=TRUE, exercise.setup="elect"}
names(dem.pres.vote.share)
names(dem.pres.vote.share) <-elect.year
names(dem.pres.vote.share)
dem.pres.vote.share
```

## Packages

Base R contains many functions to accomplish a variety of tasks. But many of the functions we will use are contained in packages that have been contributed to the Comprehensive R Archive Network (CRAN).  A package bundles together code, data, and help files. There are over 10,000 packages written for R, and more are contributed all the time. R does not automatically install or load packages in order to save memory.

Many packages will be installed for you in R Studio Cloud so that you don't have to. But if you are working on a local copy of R Studio, you will need to install them using the `install.packages()` function by passing the function the name of the package in quotes. Once installed, you load the package using the `library()` function. For example, we will use the `ggplot()` function in the  <span style="color:DarkGreen">ggplot2</span> package. You must load a package each time you start R before you can use the functions it contains (but only install it once). If you don't first load the package, you will get an error:

```{r errorlib, exercise=TRUE}
ggplot()
```
Instead, we load the package first. This makes its functions available in memory. Note the error message. Any time you receive an error that begins "could not find function" the likely problem is that you haven't loaded the packaged that contains the function!

```{r lib, exercise=TRUE}
library(ggplot2)
ggplot()
```

To get help with a package (you don't need to read the output):

```{r help1, exercise=TRUE}
help(package="ggplot2")
```

## Comments in code

If you wish to add comments to your code to help you remember why you ran a function or the arguments it takes, you place a hashtag before the comment. Anything after # on the same line will not be run.

```{r comments, exercise=TRUE}
#Sample comment
mean(dem.pres.vote.share, na.rm=TRUE) 
```

## Practice


Create a vector called **rep.pres.vote.share** containing the share of the two-party vote received by the Republican presidential candidate. Print it.  Note that you do NOT need to type in the vote shares because we are only considering votes for the major parties. That means the Republican vote share is just one minus the Democratic vote share.

```{r Repshare, exercise=TRUE}

```

```{r Repshare-hint}
Take 1-dem.pres.vote.share and assign it to rep.pres.vote.share.
```

```{r Repshare-solution}
rep.pres.vote.share <- 1-dem.pres.vote.share
rep.pres.vote.share
```

```{r Repshare-check}
grade_code(" ")
```

Create a vector called **victory.margin** that calculates the margin of victory. Here you will need to use the absolute value function, which is the `abs()` function. (You will need to list the Republican vote share first inside this function.) Print it.


```{r Repshare-me}
rep.pres.vote.share <- 1-dem.pres.vote.share
```

```{r close, exercise=TRUE, exercise.setup="Repshare-me"}

```

```{r close-hint}
The margin of victory is captured by the difference
between the Democratic vote share and the Republican vote share
and taking the absolute value of that quantity.
```

```{r close-solution}
victory.margin <- abs(rep.pres.vote.share-dem.pres.vote.share)
victory.margin
```

```{r close-check}
grade_code(" ")
```

Create a vector of Republican candidate names called **rep.names.** Print it. Just use last names. 

```{r repnames, exercise=TRUE}

```

```{r repnames-hint}
You will need to use the c() function and list, in order, the
names of the candidates in each election (in quotes), separated 
by commas. If you need help remembering the Republican 
candidates' names go back to the table earlier in the 
tutorial.
```
 
```{r repnames-solution}
rep.names <- c("Reagan", "Reagan", "Bush", "Bush", "Dole", "Bush", "Bush", "McCain", "Romney", "Trump", "Trump")
rep.names
```

```{r repnames-check}
grade_code(" ")
```

Find the mean share of the vote going to the Republican candidate.

```{r meanrep, exercise=TRUE, exercise.setup="Repshare-me"}

```

```{r meanrep-hint}
Provide the mean function the name of the vector containing
the Republican vote share.
```

```{r meanrep-solution}
mean(rep.pres.vote.share)
```

```{r meanrep-check}
grade_code(" ")
```


Find the range of the vote share received by Republican candidates.

```{r rangerep, exercise=TRUE,  exercise.setup="Repshare-me"}

```

```{r rangerep-hint}
Provide the name of the object containing the Republican vote 
share to the range() function.
```


```{r rangerep-solution}
range(rep.pres.vote.share)
```

```{r rangerep-check}
grade_code(" ")
```



